---
title: "LECTURE 5: null hypothesis testing"
subtitle: "FANR 6750 (Experimental design)"
author: "<br/><br/><br/>Fall 2023"
output:
  xaringan::moon_reader:
    css: ["default", "FANR6750.css", "FANR6750-fonts.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, echo = FALSE, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo = FALSE, fig.align = 'center', warning=FALSE, message=FALSE, fig.retina = 2)
source(here::here("R/zzz.R"))
library(emo)
library(FANR6750)
# library(gganimate)
```

class: inverse

# outline

#### 1) Sampling error and regression coefficients

<br/>  
--

#### 2) Logic of null hypothesis testing

<br/> 
--

#### 3) *t*-statistics and *t*-distribution

<br/> 

--

#### 4) *p*-values and decisions

<br/> 

--

#### 5) NHT for regression models


---
# fitting the model in `R`

```{r echo=TRUE}
data(frogdata)

fit2 <- lm(Frogs ~ Development, data = frogdata)

broom::tidy(fit2)
```


- The `(Intercept)` estimate ( $\beta_0$ ) is the expected number of frogs in a low-development plot


- The `DevelopmentHigh` estimate ( $\beta_1$ ) is the *difference* between low-development and high-development plots


- **Question**: Are there fewer frogs in high-development plots?

---

#### **Question**: Are there fewer frogs in high-development plots?

- This question is harder to answer than it might seem. Why?

--

#### Sampling error!

--

- If we did the experiment again, we'd get a different difference (maybe even a change in direction)

--

- The sample means will *always* be different (which we will prove shortly)

--

If the sample means will never be the same, we need to re-formulate our question:

- How large of a difference between our samples provides evidence that the population means are different? 


---
class:inverse,middle

# null hypothesis testing

---
# null hypothesis testing

#### How can we express our question about development and its impacts on frogs as a hypothesis?

--

If development does not influence frog abundance, $\beta_1 = 0$

- This is called the *null hypothesis* (i.e., no effect)

- Denoted $H_0$

--

If development does influence frog abundance, $\beta_1 \neq 0$

- This is the *alternative hypothesis*

- Denoted $H_a$

--

- Note that we could make the alternative more specific and say development negatively effects abundance ( $\beta_0 < 0$ )

---
# null hypothesis testing

#### Remember that $\large \beta_1$ is a **population** parameter

- We estimate population parameters from samples

--

#### What is our best estimate of $\large \beta_1$? 

--

 - The difference in sample means: $\hat{\beta}_1 = \hat{\mu}_{Low} - \hat{\mu}_{High}$
 
--

- For the frog data:

```{r echo = TRUE}
mu_high <- mean(frogs$Frogs[frogs$Development == "High"])
mu_low <- mean(frogs$Frogs[frogs$Development == "Low"])

(beta_1.hat <- mu_high - mu_low)
```

Note that this is same as the estimate of $\beta_1$ provided by the `lm` and `t-test` functions

---
# null hypothesis testing

#### Our estimate of $\large \beta_1$ is clearly not $\large 0$ but does that imply that the null hypothesis is wrong?

--

- Not necessarily! Why?

--

- Sampling error! Even if $\beta_1 = 0$, $\hat{\beta}_1$ will *never* equal 0

--

Enter *null hypothesis testing*:

> formal approach to deciding whether a statistical relationship in a sample reflects a real relationship in the population or is just due to chance 


---
# null hypothesis testing

#### NHT is based on the *expectation* of what our data should look like **if the null hypothesis is true**

--

- If our data look really different than what we expect **if the null hypothesis is true**, then it is unlikely that the null hypothesis is true and we reject $H_0$

--

- It's important to note that there is *always* a chance that our results are simply due to chance 

--

#### Type I error (i.e., false positive rate)

- The probability that we will reject $H_0$ when it is actually true

- Usually denoted $\alpha$

- Generally, $\alpha = 0.05$ or $\alpha = 0.01$ are accepted Type I error rates

---
# null hypothesis testing

#### How do we know what the data should look like **if the null hypothesis is true**?

--

- Again, we can use theory to tell us about long-term expectations (similar to the sampling distribution)

--

- But maybe easier to understand using simulation

---
# simulating data under the null hypothesis

#### Imagine the null hypothesis is true in the measurement error example (i.e., measurement error = 0)

- Any difference between the sample mean and 0 is just due to sampling error

--

- We can simulate a "sample" in `R` using the `rnorm()` function

```{r echo = TRUE}
y <- rnorm(n = 10, mean = 0, sd = 1)

mean(y) - 0
```


--

Note that the we fixed the mean of the normal distribution to be 0, so we know the null hypothesis is true

---
# simulating data under the null hypothesis

Instead of just taking the difference $\bar{y} - 0$, we can standardize the differences by dividing by the standard error:

```{r echo = TRUE}
se.y <- sd(y)/sqrt(10)
(mean(y) - 0)/se.y
```

--

- This tells us how many standard errors away from 0 the sample mean is 

- We'll call this value $t$

$$\large t = \frac{\bar{y} - 0}{SE_y}$$


---
# simulating data under the null hypothesis

#### Generate and plot 1000 sample means:

```{r echo = TRUE}
t <- numeric(length = 1000)

for(i in 1:1000){
  y <-  rnorm(10, mean = 0, sd = 1)
  t[i] <- mean(y)/(sd(y)/sqrt(10))
}
```

```{r fig.width=6, fig.height=2.5}
ggplot(data.frame(t = t), aes(x = t)) +
  geom_histogram(color = "grey75") +
  geom_rug(alpha = 0.1) +
  scale_x_continuous("t") +
  scale_y_continuous("Count")
```

---
# simulating data under the null hypothesis

```{r fig.width=6, fig.height=3}
ggplot(data.frame(t = t), aes(x = t)) +
  geom_histogram(color = "grey75") +
  geom_rug(alpha = 0.1) +
  scale_x_continuous("t") +
  scale_y_continuous("Count")
```

#### Note that:

--
1. The null hypothesis was true for every sample

--

2. Some sample means were bigger than expected, some were smaller (all due to sampling error!)

--

3. The resulting distribution looks *kind of* normal

---
## THE *t*-DISTRIBUTION

The distribution of the *t*-statistics is not quite normally distributed

--

Instead, theory says that the *t*-statistics will follow a $t$-distribution with $n - 1$ degrees of freedom (**if the null hypothesis is true**!)

```{r t, fig.width=8, fig.height=6}
curve(dt(x, 1), -5, 5, ylim=c(0, 0.5),
      xlab="t value", ylab="Probability density",
      cex.lab=1.5, lwd=1.5)
curve(dt(x, 5), -5, 5, add=TRUE, col="blue", lwd=1.5)
curve(dnorm(x, 0, 1), -5, 5, add=TRUE,  lty=3, lwd=1.5)
legend(-5, 0.5, c("Standard normal distribution", "t distribution (df=5)", "t distribution (df=1)"),
       lty=c(3, 1, 1), col=c("black", "blue", "black"), lwd=1.5)
```

---
## THE *t*-DISTRIBUTION

#### More about the *t*-distribution

--

- Continuous probability distribution

--

- Symmetrical with mean = 0

--

- More mass in the tails as degrees of freedom get smaller (i.e., more extreme values become more likely)


--

- For sample sizes $n \gt 30$, the $t$-distribution is essentially a standard normal distribution with mean = 0 and SD = 1

--

- Published by William Sealy Gosset in 1908 under the pseudonym "Student". Gosset worked for Guinness and was interested in quality control of beer ingredients 

---
# null hypothesis testing

#### Quick review:

--

1. The null hypothesis $\large H_0$ is that there is no effect or no difference

--

2. The *t*-statistic measures the difference between the sample mean and its hypothesized value (under the null) relative to its standard error 

--

3. **If the null hypothesis is true**, the *t*-statistics from repeated samples follow a *t*-distribution with $\large n-1$ degrees of freedom

--

Importantly, because we can quantify properties of the *t*-distribution, we can compare the *t*-statistic calculated from our observed sample to the expected values under the null hypothesis

--
- If our observed *t*-statistic would be unlikely under the null hypothesis, we can conclude that the null hypothesis is false

--

- This is the logic behind null hypothesis testing

---
# null hypothesis testing

---
# null hypothesis testing

#### Is there evidence to reject $\large H_0$?

```{r fig.width=6, fig.height=4}
df <- data.frame(t = t,
                 group = ifelse(t > -1.07 & t < 1.07, "0", "1"))

ggplot() +
  geom_histogram(data = df, aes(x = t), 
                 alpha = 0.75, bins = 100) +
  geom_vline(xintercept = 1.05, color = "#D47500", linetype = "dashed", size = 1) +
  geom_vline(xintercept = mean(t), color = "black") +
  scale_y_continuous("") + theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())
```

---
# null hypothesis testing

#### Measurement error example

```{r echo =TRUE}
y <- c(-0.062, -0.38, 0.85, -0.58,  0.53,  0.09,  0.31,  0.77,  0.59,  -0.17)

(mean.y <- mean(y))

(se.y <- sd(y)/sqrt(10))

(t.stat <- mean.y / se.y)
```

--

#### Is there evidence to reject $\large H_0$?

---
# null hypothesis testing

#### Measurement error example


```{r fig.width=6, fig.height=4}
df <- data.frame(t = t,
                 group = ifelse(t > -t.stat & t < t.stat, "0", "1"))

ggplot() +
  geom_histogram(data = df, aes(x = t, fill = as.factor(group)), 
                 alpha = 0.75, bins = 100) +
  geom_vline(xintercept = t.stat, color = "#D47500", linetype = "dashed", size = 1) +
  geom_vline(xintercept = mean(t), color = "black") +
  scale_y_continuous("") + theme(axis.text.y = element_blank(), axis.ticks.y = element_blank()) +
  guides(fill = "none") +
  labs(subtitle = "Distribution of 1000 t-statistics")
```

Approximately 18% of the simulated values of $t$ are larger than 1.25 (or smaller than -1.25)

- Put another way, **if the null hypothesis is true**, there is about a 1 in 5 chance of observing $t \geq 1.25$ 

---
# null hypothesis testing

In reality, we don't need to simulate the distribution of $t$ every time we do an experiment

<br/>
```{r p, fig.width=5, fig.height=3}
op <- par(mai=c(0.8, 0.2, 0.2, 0.2))
curve(dt(x, df=9), -4, 4, xlab="t value", ylab="", yaxt="n",
      ylim=c(0,0.5),
      frame=FALSE , main="t-distribution with df=9", cex.main=1.5)
xs1 <- seq(qt(.025, df=9), -4, by=-0.1)
ys1 <- dt(xs1, df=9)
xs2 <- seq(qt(.975, df=9), 4, by=0.1)
ys2 <- dt(xs2, df=9)
xs3 <- seq(qt(pt(-1.251,9), df=9), -4, by=-0.1)
ys3 <- dt(xs3, df=9)
xs4 <- seq(qt(pt(1.251,9), df=9), 4, by=0.1)
ys4 <- dt(xs4, df=9)
polygon(c(xs3, rev(xs3)), c(rep(0, length(xs3)), rev(ys3)), col=gray(0.7))
polygon(c(xs4, rev(xs4)), c(rep(0, length(xs4)), rev(ys4)), col=gray(0.7))
#polygon(c(xs1, rev(xs1)), c(rep(0, length(xs1)), rev(ys1)), col="red")
#polygon(c(xs2, rev(xs2)), c(rep(0, length(xs2)), rev(ys2)), col="red")
text(-1.251, dt(0,9), "t=-1.251", pos=3)
text(1.251, dt(0,9), "t=1.251", pos=3)
#text(-3, dt(0,18), "t=-3", pos=3)
#text(3, dt(0,18), "t=3", pos=3)
arrows(-1.07, dt(0,9), -1.251, dt(-1.07,9), length=0.1)
arrows(1.07, dt(0,9), 1.251, dt(1.07,9), length=0.1)
par(op)
```

--

- Easy (in `R`) to calculate the area of grey shaded regions, which is the probability of getting a value $t$ larger in magnitude than 1.251 **if the null hypothesis true**

--

- This is called a *p*-value

--

- In our example, $p = 0.174$. Would you reject the null?

---
# more on *p*-values

#### The *p* -value tells you how likely your observations (or more extreme) would be **if the null hypothesis is true**  

--

#### A *p* -value does not tell us how much evidence there is in favor of a particular difference in means

--

#### What factors result in a small *p* -value?

--

- The sample mean is far from 0

- And/or the SE is small

---
# *p*-values and type i error

#### In NHST, our conclusion must be to either reject or "fail to reject" the null hypothesis

--

#### When we reject the null hypothesis, there is always a chance that we do so mistakenly

- Due to sampling error, there is always a chance we get a large value of *t* even if the null hypothesis is true

- These "false positive" mistakes are referred to as *Type I error* (denoted $\alpha$)

--

#### The probability of type I error is measured by the *p*-value

--

#### Generally, we want to avoid false positive conclusions. Why?

--

- Type I error rates of $\lt 5$% or $\lt 1$% are generally considered reasonable


---
# critical values

#### Before statistical software made it easy to calculate *p*-values, researchers would look up *critical values*

- For a given sample size (degrees of freedom) and $\alpha$, what is the associated value of *t*

- If your calculated *t* is $\geq$ the critical value, reject the null hypothesis

```{r cv, fig.width=6, fig.height=4}
op <- par(mai=c(0.8, 0.2, 0.2, 0.2))
curve(dt(x, df=9), -4, 4, xlab="t value", ylab="", yaxt="n",
      ylim=c(0,0.5),
      frame=FALSE , main="t distribution with df=9", cex.main=1.5)
xs1 <- seq(qt(.025, df=9), -4, by=-0.1)
ys1 <- dt(xs1, df=9)
xs2 <- seq(qt(.975, df=9), 4, by=0.1)
ys2 <- dt(xs2, df=9)
polygon(c(xs1, rev(xs1)), c(rep(0, length(xs1)), rev(ys1)), col="red")
polygon(c(xs2, rev(xs2)), c(rep(0, length(xs2)), rev(ys2)), col="red")
text(xs1[1], dt(0,9)/1.5, "critical value\nt=-2.26", pos=3)
text(xs2[1], dt(0,9)/1.5, "critical value\nt=2.26", pos=3)
arrows(xs1[1], dt(0,9)/1.5, xs1[1], ys1[1], length=0.1)
arrows(xs2[1], dt(0,9)/1.5, xs2[1], ys2[1], length=0.1)
#text(-3, dt(0,18), "t=-3", pos=3)
#text(3, dt(0,18), "t=3", pos=3)
#arrows(-3, dt(0,18), -3, dt(-3,18), length=0.1)
#arrows(3, dt(0,18), 3, dt(3,18), length=0.1)
par(op)

```

---
# frog example

#### Back to the frog example

```{r echo=TRUE}
data(frogdata)

fit2 <- lm(Frogs ~ Development, data = frogdata)

broom::tidy(fit2)
```

--

Now we're ready to interpret the rest of this output

---
# nhst for regression parameters

#### It turns out, regression parameters also follow a *t*-distribution **if the null hypothesis is true**

--

#### In a regression context:

$$\large t = \frac{\hat{\beta}_1 - \beta_1}{SE(\hat{\beta}_1)}$$

--

#### For the frog example:

```{r echo = TRUE}
beta1.hat <- mu_high - mu_low
(t <- (beta1.hat - 0)/2.64)
```

---
# nhst for regression parameters

```{r echo=TRUE}
broom::tidy(fit2)
```

Notice that the `lm` output provides a *t*-statistic and *p*-value for both the slope coefficient and the intercept

- How do we interpret the intercept *p*-value? 

- Is the *p*-value for the intercept biologically meaningful?


---
# nhst for t-tests

As we saw before, the frog model can be fit as a two-sample *t*-test

```{r echo = TRUE}
t.test(Frogs ~ Development, data = frogs)
```

--

Notice that the *t*-statistic and *p*-value is the same as the `lm` model

---
# looking ahead

<br/>


### **Next time**: Statistical power

