---
title: "LECTURE 18: logistic regression"
subtitle: "FANR 6750 (Experimental design)"
author: "<br/><br/><br/>Fall 2023"
output:
  xaringan::moon_reader:
    css: ["default", "FANR6750.css", "FANR6750-fonts.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, echo = FALSE, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo = FALSE, fig.align = 'center', warning=FALSE, message=FALSE, fig.retina = 2)
source(here::here("R/zzz.R"))
library(emo)
library(FANR6750)
library(dplyr)
library(kableExtra)
# library(gganimate)
```

# outline 


1) Motivation

<br/>

--

2) Model

<br/>

--

3) Model fitting

<br/>
--


4) Model interpretation


---
# logistic regression

Logistic regression is a specific type of GLM in which the response variable follows a binomial distribution and the link function is (usually) the logit  

--

Logistic regression is used to model a binary response variable as a function of explanatory variables

--

#### Examples:

- Is presence of a focal species related to habitat type?  

- Is survival a function of body condition?  

- Is disease prevalence related to population density?  


---
# logistic regression

<br/>

$$\large y_i \sim Bernoulli(p_i)$$

$$\large log\bigg(\frac{p_i}{1 - p_i}\bigg) = logit(p_i) = \beta_0 + \beta_1x_{i1} + \beta_2x_{i2} + ...$$

--

#### where:

- $\large y_i$ is the response (coded as 0/1)  

- $\large p_i$ is the probability of success for sample unit *i*

- $\large \frac{p_i}{1 - p_i}$ is the *odds* of success

- $\large log\big(\frac{p_i}{1 - p_i}\big)$ is the log odds 

---
# logit link

The logit link  allows us to transform the probability $p_i$ (constrained to $0$ to $1$) to the real scale ( $-\infty$ to $\infty$)

$$\large \log\bigg(\frac{0.5}{1-0.5}\bigg) = log(1) = 0$$
--

The inverse logit $\frac{e^\mu}{1+e^\mu}$ allows us to transform the linear predictor to the probability scale

$$\large \frac{e^0}{1+e^0} = \frac{1}{2}$$

--

The odds ratio $\frac{p}{1-p}$ varies from $0$ to $\infty$ and indicate the probability of success relative to the probability of failure

--

- $1:2$ odds = $0.33/(1 - 0.33)$ = failure is twice as likely as success

- $4:1$ odds = $0.8/(1 - 0.8)$ = success if four times a likely as failure

---
# logit link

<br/>

```{r}
logit_df <- data.frame(metric = c("log odds = log(p/(1-p))", "odds = p/(1-p)", "p"),
                       v1 = c(round(qlogis(0.1), 2), 0.1/(1-0.1), 0.1),
                       v2 = c(round(qlogis(0.25), 2), 0.25/(1-0.25), 0.25),
                       v3 = c(round(qlogis(0.5), 2), 0.5/(1-0.5), 0.5),
                       v4 = c(round(qlogis(0.8), 2), 0.8/(1-0.8), 0.8),
                       v5 = c(round(qlogis(0.99), 2), 0.99/(1-0.99), 0.99))

logit_df %>% 
  kable(digits = 2, col.names = NULL)
```

--

<br/>
```{r fig.height=3.5, fig.width=10}
x <- seq(-5, 5, by = 0.1)
logit_df2 <- data.frame(mu = c(-1 * x, 0.5*x, 2*x),
                        p = c(plogis(-1 * x), plogis(0.5*x), plogis(2*x)),
                        x = x,
                        beta = factor(rep(c("-1", "0.5", "2"), each = length(x)),
                                      levels = c("-1", "0.5", "2")))
r <- ggplot(logit_df2, aes(x = x, y = mu, color = beta)) +
  geom_path() +
  scale_x_continuous("x") +
  scale_y_continuous(expression(mu)) +
  labs(subtitle = expression(paste("0 + ", beta[1], "*x"))) +
    scale_color_discrete(labels = c(expression(paste(beta[1], "= -1")), 
                                  expression(paste(beta[1], "= 0.5")), 
                                  expression(paste(beta[1], "= 2")))) +
  guides(color = "none")

q <- ggplot(logit_df2, aes(x = x, y = p, color = beta)) +
  geom_path() +
  scale_x_continuous("x") +
  scale_y_continuous("p") +
  labs(subtitle = expression(paste("ilogit(0 + ", beta[1], "*x)"))) +
  scale_color_discrete(labels = c(expression(paste(beta[1], "= -1")), 
                                  expression(paste(beta[1], "= 0.5")), 
                                  expression(paste(beta[1], "= 2"))))
cowplot::plot_grid(r, q, align = "hv", nrow = 1, rel_widths = c(0.425, 0.575))
```


---
# example

Imagine we are interested in the effects of elevation and habitat on the probability of occurrence for a rare orchid

```{r echo = TRUE}
data("orchidata")
head(orchiddata, n=12)
```

---
# raw data

```{r echo = TRUE, fig.height=5, fig.width=7}
ggplot(orchiddata, aes(x = elevation, y = presence)) +
  geom_point() +
  scale_y_continuous("Orchid Occurrence") + scale_x_continuous("Elevation")
```

---
# raw data

```{r echo = TRUE, fig.height=4, fig.width=7}
library(dplyr)

orchiddata %>% group_by(habitat) %>% summarise(group.prob = mean(presence)) %>%
  ggplot(., aes(x = habitat, y = group.prob)) +
  geom_col(fill = "grey70", color = "black") +
  scale_y_continuous("Proportion of sites with orchids") + scale_x_discrete("Habitat")
```

---
# the `glm` function

<br/>

```{r echo = TRUE, eval = FALSE}
fm1 <- glm(presence ~ habitat + elevation, 
           family=binomial(link="logit"), 
           data = orchiddata)
broom::tidy(fm1)
```

<br/>

```{r echo = FALSE, eval = TRUE}
fm1 <- glm(presence ~ habitat + elevation, family=binomial(link="logit"), data = orchiddata)
kable(broom::tidy(fm1))
```

---
# intepreting parameters

```{r echo = FALSE, eval = TRUE}
fm1 <- glm(presence ~ habitat + elevation, family=binomial(link="logit"), data = orchiddata)
kable(broom::tidy(fm1)) %>%
  kable_styling(font_size = 10)
```

--

- The probability of occurrence in maple-dominated forest at sea level = `r round(plogis(coef(fm1)[1]), 2)` (`plogis(-0.996)`)

--

- The probability of occurrence in oak-dominated forest at sea level = `r round(plogis(coef(fm1)[1] + coef(fm1)[2]), 2)` (`plogis(-0.996 - 0.09)`)

--

    + The *odds* of occurrence in oak (relative to maple) are `exp(-0.09)` = `r round(exp(-0.09), 2)`
    
--

- The probability of occurrence in pine-dominated forest at sea level = `r round(plogis(coef(fm1)[1] + coef(fm1)[3]), 2)` (`plogis(-0.996 - 0.34)`)

--

    + The *odds* of occurrence in pine (relative to maple) are `exp(-0.34)` = `r round(exp(-0.34), 2)`
    
---
# intepreting parameters

```{r echo = FALSE, eval = TRUE}
fm1 <- glm(presence ~ habitat + elevation, family=binomial(link="logit"), data = orchiddata)
kable(broom::tidy(fm1)) %>%
  kable_styling(font_size = 10)
```

--

What about elevation? 

--

First, let's visualize the fitted relationship

---
# example

<br/>

```{r fig.height=6, fig.width=8}
newdat <- data.frame(elevation=seq(12, 489, length = 50), 
                     habitat="Oak")
pred.link <- predict(fm1, newdata=newdat, se.fit=TRUE, type="link")
newdat$mu <- plogis(pred.link$fit)
newdat$lower <- plogis(pred.link$fit - 1.96*pred.link$se.fit)
newdat$upper <- plogis(pred.link$fit + 1.96*pred.link$se.fit)

ggplot(orchiddata, aes(x = elevation, y = presence)) +
  geom_point() +
  scale_y_continuous("Orchid Occurrence") + 
  scale_x_continuous("Elevation") +
  geom_line(data = newdat, aes(x = elevation, y = mu)) +
  geom_line(data = newdat, aes(x = elevation, y = lower), linetype = "longdash") +
  geom_line(data = newdat, aes(x = elevation, y = upper), linetype = "longdash")
```

---
# example

What's the change in probability of occurrence at 1m vs 0m elevation (in maple habitat)?

- `plogis(-0.996 + 0.014 * 1) - plogis(-0.996 + 0.014 * 0)` = `r round(plogis(coef(fm1)[1] + coef(fm1)[4] * 1), 3) - round(plogis(coef(fm1)[1]), 3)`

--

What's the change in probability of occurrence at 101m vs 100m elevation ?

- `plogis(-0.996 + 0.014 * 101) - plogis(-0.996 + 0.014 * 100)` = `r round(plogis(coef(fm1)[1] + coef(fm1)[4] * 101), 3) - round(plogis(coef(fm1)[1]+ coef(fm1)[4] * 100), 3)`

--

Change in probability is not linear

```{r fig.height=3.5, fig.width=8}
ggplot(orchiddata, aes(x = elevation, y = presence)) +
    geom_point() +
    scale_y_continuous("Orchid Occurrence") + 
    scale_x_continuous("Elevation", breaks = c(50, 100, 350, 400)) +
    geom_line(data = newdat, aes(x = elevation, y = mu)) +
    geom_segment(aes(x= 100, xend = 100, y = -Inf, yend = plogis(coef(fm1)[1] + coef(fm1)[4]*93)), 
                 linetype = "longdash", color = "grey25", linewidth = 0.25) +
    geom_segment(aes(x= -Inf, xend = 100, y =  plogis(coef(fm1)[1] + coef(fm1)[4]*93), 
                     yend = plogis(coef(fm1)[1] + coef(fm1)[4]*93)), 
                 linetype = "longdash", color = "grey25", linewidth = 0.25) +
    geom_segment(aes(x= 50, xend = 50, y = -Inf, yend = plogis(coef(fm1)[1] + coef(fm1)[4]*43)), 
                 linetype = "longdash", color = "grey25", linewidth = 0.25) +
    geom_segment(aes(x= -Inf, xend = 50, y =  plogis(coef(fm1)[1] + coef(fm1)[4]*43), 
                     yend = plogis(coef(fm1)[1] + coef(fm1)[4]*43)),
                 linetype = "longdash", color = "grey25", linewidth = 0.25) +
    geom_segment(aes(x= 400, xend = 400, y = -Inf, yend = plogis(coef(fm1)[1] + coef(fm1)[4]*393)),
                 linetype = "longdash", color = "grey25", linewidth = 0.25) +
    geom_segment(aes(x= -Inf, xend = 400, y =  plogis(coef(fm1)[1] + coef(fm1)[4]*393), 
                     yend = plogis(coef(fm1)[1] + coef(fm1)[4]*393)),
                 linetype = "longdash", color = "grey25", linewidth = 0.25) +
    geom_segment(aes(x= 350, xend = 350, y = -Inf, yend = plogis(coef(fm1)[1] + coef(fm1)[4]*343)),
                 linetype = "longdash", color = "grey25", linewidth = 0.25) +
    geom_segment(aes(x= -Inf, xend = 350, y =  plogis(coef(fm1)[1] + coef(fm1)[4]*343), 
                     yend = plogis(coef(fm1)[1] + coef(fm1)[4]*343)), 
                 linetype = "longdash", color = "grey25", linewidth = 0.25)

```


---
# example

Change in probability is not linear. 

--

But the change is *odds* is 

- Odds ratio: $\frac{p_2}{1 - p_2}/\frac{p_1}{1 - p_1}$

.small-code[
```{r echo = TRUE}
p1 <- plogis(-0.996 + 0.014 * 0)
p2 <- plogis(-0.996 + 0.014 * 1)
(OR1 <- (p2/(1-p2))/(p1/(1-p1)))

p3 <- plogis(-0.996 + 0.014 * 400)
p4 <- plogis(-0.996 + 0.014 * 401)
(OR2 <- (p4/(1-p4))/(p3/(1-p3)))

```
]

--

What is the change in odds for one unit change in elevation?

--

- $e^{\beta_{Elev}}$ = `exp(0.014)` = `r exp(0.014)`

---
# summary

Logistic regression is used when we want to model a binary response as a function of explanatory variables

--

The response $y_i$ is modeled as arising from a Bernoulli distribution with probability $p_i$ and the logit link is used to map the linear predictor to the probability scale

--

All of the linear modeling concepts we have learned this semester (continuous/categorical explanatory variables, multiple regression, interactions, random effects) can be used within the logistic regression framework

--

Parameter estimates from the logistic regression measure the change in log odds for one unit change in the explanatory variables

  - Log odds are constant across all values of $x$, but the probabilities are not  


---
# looking ahead

<br/>

### **Next time**: Poisson regression

<br/>

### **Reading**: [Fieberg chp. 15](https://statistics4ecologists-v1.netlify.app/poissonr)


