---
title: "LECTURE 6: statistical power"
subtitle: "FANR 6750 (Experimental design)"
author: "<br/><br/><br/>Fall 2024"
output:
  xaringan::moon_reader:
    css: ["default", "FANR6750.css", "FANR6750-fonts.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, echo = FALSE, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo = FALSE, fig.align = 'center', warning=FALSE, message=FALSE, fig.retina = 2)
source(here::here("R/zzz.R"))
library(emo)
library(FANR6750)
# library(gganimate)
```

# outline 

<br/>

1) Motivation

<br/>

--

2) Type I and Type II error

<br/>

--

3) Two-sample *t*-test

<br/>

--

4) ANOVA

---
class: inverse

# motivation

<br/>
<br/>

> A statistical test will not be able to detect a true difference if the sample size is too small compared with the magnitude of the difference.

> Since data are sampled at random, there is always a risk of reaching a wrong conclusion, and things can go wrong in two ways - Dalgaard (2008)

---
# type i & type ii errors

#### Type I error (i.e., false positive)

> The null hypothesis is correct, but the test rejects it.

$$\large \alpha = Pr(Type\;I\;error)$$

--

#### Type II error (i.e., false negative)

> The null hypothesis is wrong, but the test fails to reject it.

$$\large \beta = Pr(Type\;II\;error)$$

--

#### Power

> The test's ability to reject a false null hypothesis.

$$\large Power = 1 - \beta$$

---
# type i & type ii errors

#### The type I error rate is set by the scientist

--

#### The type II error rate, and hence the power of the test, depends on many factors  

--

#### In the context of a linear model, these are:

1) Magnitude of the slope coefficients ( $\beta$ )  

2) Standard deviation (or variance) of population ( $\sigma$ )  

3) The sample size ( $n$ )  

4) The Type I error rate ( $\alpha$ )  

---
# magnitude of the difference

```{r echo = FALSE, fig.align="c"}
x <- seq(from = 0, to = 200, length.out = 200)

delta_df <- data.frame(x = x, 
                       y1 = c(dnorm(x, 98, 10), dnorm(x, 102, 10)),
                       y2 = c(dnorm(x, 80, 10), dnorm(x, 120, 10)),
                       y3 = c(dnorm(x, 80, 40), dnorm(x, 120, 40)),
                       population = rep(c("A", "B"), each = 200))

ggplot(delta_df, aes(x = x, y = y1, color = population)) +
  geom_path(size = 1.75) +
  scale_x_continuous("") +
  scale_y_continuous("") +
  guides(color = "none") +
  labs(subtitle = "Hard to detect a difference")
```

---
# magnitude of the difference

```{r echo = FALSE, fig.align="c"}

ggplot(delta_df, aes(x = x, y = y2, color = population)) +
  geom_path(size = 1.75) +
  scale_x_continuous("") +
  scale_y_continuous("") +
  guides(color = "none") +
  labs(subtitle = "Easier to detect a difference")
```

---
# standard deviation

```{r echo = FALSE, fig.align="c"}

ggplot(delta_df, aes(x = x, y = y2, color = population)) +
  geom_path(size = 1.75) +
  scale_x_continuous("") +
  scale_y_continuous("") +
  guides(color = "none") +
  labs(subtitle = "Easier to detect a difference")
```


---
# standard deviation

```{r echo = FALSE, fig.align="c"}

ggplot(delta_df, aes(x = x, y = y3, color = population)) +
  geom_path(size = 1.75) +
  scale_x_continuous("") +
  scale_y_continuous("") +
  guides(color = "none") +
  labs(subtitle = "Hard to detect a difference")
```

---
# sample size

.pull-left[
```{r echo = FALSE, fig.align="c"}
delta_df <- data.frame(x = x, 
                       y1 = c(dnorm(x, 90, 10), dnorm(x, 100, 10)),
                       population = rep(c("A", "B"), each = 200))

n1 <- data.frame(x1 = c(rnorm(10, 90, 10), rnorm(10, 100, 10)),
                # x2 = c(rnorm(100, 97, 10), rnorm(100, 103, 10)),
                y = rep(c(0, 0.001), each = 10),
                population = rep(c("A", "B"), each = 10))

ggplot(delta_df, aes(x = x, y = y1, color = population)) +
  geom_path(size = 1.75) +
  geom_point(data = n1, aes(x = x1, y = y, color = population), alpha = 0.5) +
  scale_x_continuous("") +
  scale_y_continuous("") +
  guides(color = "none") +
  labs(subtitle = "Hard to detect a difference (n = 10)")
```
]

.pull-right[
```{r echo = FALSE, fig.align="c"}

ggplot() +
  geom_boxplot(data = n1, aes(y = x1, x = y, color = population), alpha = 0.5) +
  scale_x_continuous("") +
  scale_y_continuous("") +
  guides(color = "none") 
```
]

---
# sample size

.pull-left[
```{r echo = FALSE, fig.align="c"}
n2 <- data.frame(x1 = c(rnorm(1000, 90, 10), rnorm(1000, 100, 10)),
                y = rep(c(0, 0.001), each = 1000),
                population = rep(c("A", "B"), each = 1000))

ggplot(delta_df, aes(x = x, y = y1, color = population)) +
  geom_path(size = 1.75) +
  geom_point(data = n2, aes(x = x1, y = y, color = population), alpha = 0.25) +
  scale_x_continuous("") +
  scale_y_continuous("") +
  guides(color = "none") +
  labs(subtitle = "Easier to detect a difference (n = 1000)")
```
]

.pull-right[
```{r echo = FALSE, fig.align="c"}

ggplot() +
  geom_boxplot(data = n2, aes(y = x1, x = y, color = population), alpha = 0.5) +
  scale_x_continuous("") +
  scale_y_continuous("") +
  guides(color = "none") 
```
]

---
# type i error rate

#### $\large \alpha = 0.05$

```{r cv, fig.height = 6, fig.width=8}
ggplot(NULL, aes(c(-5,5))) +
  geom_area(stat = "function", fun = dnorm, fill = "#00998a", xlim = c(-5, qnorm(0.025, 0, 1))) +
  geom_area(stat = "function", fun = dnorm, fill = "grey80", xlim = c(qnorm(0.025, 0, 1), qnorm(0.975, 0, 1))) +
  geom_area(stat = "function", fun = dnorm, fill = "#00998a", xlim = c(qnorm(0.975, 0, 1), 5)) +
  labs(subtitle = "Easier to detect difference", x = "t", y = "") +
  scale_y_continuous(breaks = NULL) +
  scale_x_continuous(breaks = c(qnorm(0.025, 0, 1), qnorm(0.975, 0, 1)))
```

---
# type i error rate

#### $\large \alpha = 0.001$

```{r cv2, fig.height = 6, fig.width=8}
ggplot(NULL, aes(c(-5,5))) +
  geom_area(stat = "function", fun = dnorm, fill = "#00998a", xlim = c(-5, qnorm(0.005, 0, 1))) +
  geom_area(stat = "function", fun = dnorm, fill = "grey80", xlim = c(qnorm(0.005, 0, 1), qnorm(0.995, 0, 1))) +
  geom_area(stat = "function", fun = dnorm, fill = "#00998a", xlim = c(qnorm(0.995, 0, 1), 5)) +
  labs(subtitle = "Harder to detect difference", x = "t", y = "") +
  scale_y_continuous(breaks = NULL) +
  scale_x_continuous(breaks = c(qnorm(0.005, 0, 1), qnorm(0.995, 0, 1)))
```

---
# factors affecting power

#### In general, power increases when:

1) The difference in means/magnitude of slope increases  

2) The standard deviation **of the population** decreases  

3) The sample size increases  

4) The Type I error rate increases  

--

#### Which of these, as researchers, do we have control over?

---
# example in `R`

.pull-left[

#### $\large \mu_1 = 90$, $\large \mu_2 = 100$

#### $\large \sigma = 5$

```{r}
delta_df <- data.frame(x = x, 
                       y1 = c(dnorm(x, 90, 5), dnorm(x, 100, 5)),
                       population = rep(c("A", "B"), each = 200))

n2 <- data.frame(x1 = c(rnorm(5, 90, 5), rnorm(5, 100, 5)),
                 y = rep(c(0, 0.001), each = 5),
                 population = rep(c("A", "B"), each = 5))
mu_df <- data.frame(population = c("A", "B"),
                    mu = c(90, 100))

ggplot(delta_df, aes(x = x, y = y1, color = population)) +
  geom_path(size = 1.75) +
  # geom_point(data = n2, aes(x = x1, y = y, color = population), alpha = 0.75) +
  geom_vline(data = mu_df, aes(xintercept = mu, color = population), 
             linetype = "dashed", size = 1) +
  scale_y_continuous("") +
  guides(color = "none") +
  scale_x_continuous("", limits = c(70, 120), breaks = c(90, 100)) +
  theme(axis.text.x = element_text(size = 20))
```

]

---
# example in `R`

.pull-left[

#### $\large \mu_1 = 90$, $\large \mu_2 = 100$

#### $\large \sigma = 5$

```{r}
delta_df <- data.frame(x = x, 
                       y1 = c(dnorm(x, 90, 5), dnorm(x, 100, 5)),
                       population = rep(c("A", "B"), each = 200))

n2 <- data.frame(x1 = c(rnorm(5, 90, 5), rnorm(5, 100, 5)),
                 y = rep(c(0, 0.001), each = 5),
                 population = rep(c("A", "B"), each = 5))
mu_df <- data.frame(population = c("A", "B"),
                    mu = c(90, 100))

ggplot(delta_df, aes(x = x, y = y1, color = population)) +
  geom_path(size = 1.75) +
  geom_point(data = n2, aes(x = x1, y = y, color = population), alpha = 0.75) +
  geom_vline(data = mu_df, aes(xintercept = mu, color = population), 
             linetype = "dashed", size = 1) +
  scale_y_continuous("") +
  guides(color = "none") +
  scale_x_continuous("", limits = c(70, 120), breaks = c(90, 100)) +
  theme(axis.text.x = element_text(size = 20))
```
]

--

.pull-right[
#### $\large n = 5$
.small-code[
```{r echo = TRUE}
power.t.test(n = 5, 
             delta = 10, 
             sd = 5, 
             sig.level = 0.05, 
             power = NULL)
```
]
]

---
# example in `R`

.pull-left[

#### $\large \mu_1 = 90$, $\large \mu_2 = 100$

#### $\large \sigma = 5$

```{r}
delta_df <- data.frame(x = x, 
                       y1 = c(dnorm(x, 90, 5), dnorm(x, 100, 5)),
                       population = rep(c("A", "B"), each = 200))

n2 <- data.frame(x1 = c(rnorm(15, 90, 5), rnorm(15, 100, 5)),
                 y = rep(c(0, 0.001), each = 15),
                 population = rep(c("A", "B"), each = 15))
mu_df <- data.frame(population = c("A", "B"),
                    mu = c(90, 100))

ggplot(delta_df, aes(x = x, y = y1, color = population)) +
  geom_path(size = 1.75) +
  geom_point(data = n2, aes(x = x1, y = y, color = population), alpha = 0.75) +
  geom_vline(data = mu_df, aes(xintercept = mu, color = population), 
             linetype = "dashed", size = 1) +
  scale_y_continuous("") +
  guides(color = "none") +
  scale_x_continuous("", limits = c(70, 120), breaks = c(90, 100)) +
  theme(axis.text.x = element_text(size = 20))
```
]

.pull-right[
#### $\large n = 15$
.small-code[
```{r echo = TRUE}
power.t.test(n = 15, 
             delta = 10, 
             sd = 5, 
             sig.level = 0.05, 
             power = NULL)
```
]
]

---
# example in `R`

.pull-left[

#### $\large \mu_1 = 90$, $\large \mu_2 = 100$

#### $\large \sigma = 5$

```{r}
delta_df <- data.frame(x = x, 
                       y1 = c(dnorm(x, 90, 5), dnorm(x, 100, 5)),
                       population = rep(c("A", "B"), each = 200))

n2 <- data.frame(x1 = c(rnorm(15, 90, 5), rnorm(15, 100, 5)),
                 y = rep(c(0, 0.001), each = 15),
                 population = rep(c("A", "B"), each = 15))
mu_df <- data.frame(population = c("A", "B"),
                    mu = c(90, 100))

ggplot(delta_df, aes(x = x, y = y1, color = population)) +
  geom_path(size = 1.75) +
  geom_point(data = n2, aes(x = x1, y = y, color = population), alpha = 0.75) +
  geom_vline(data = mu_df, aes(xintercept = mu, color = population), 
             linetype = "dashed", size = 1) +
  scale_y_continuous("") +
  guides(color = "none") +
  scale_x_continuous("", limits = c(70, 120), breaks = c(90, 100)) +
  theme(axis.text.x = element_text(size = 20))
```
]

.pull-right[
#### $\large \alpha = 0.001$
.small-code[
```{r echo = TRUE}
power.t.test(n = 15, 
             delta = 10, 
             sd = 5, 
             sig.level = 0.001, 
             power = NULL)
```
]
]

---
# example in `R`

.pull-left[

#### $\large \mu_1 = 94$, $\large \mu_2 = 97$

#### $\large \sigma = 5$

```{r}
delta_df <- data.frame(x = x, 
                       y1 = c(dnorm(x, 94, 5), dnorm(x, 97, 5)),
                       population = rep(c("A", "B"), each = 200))

n2 <- data.frame(x1 = c(rnorm(15, 94, 5), rnorm(15, 97, 5)),
                 y = rep(c(0, 0.001), each = 15),
                 population = rep(c("A", "B"), each = 15))
mu_df <- data.frame(population = c("A", "B"),
                    mu = c(94, 97))

ggplot(delta_df, aes(x = x, y = y1, color = population)) +
  geom_path(size = 1.75) +
  geom_point(data = n2, aes(x = x1, y = y, color = population), alpha = 0.75) +
  geom_vline(data = mu_df, aes(xintercept = mu, color = population), 
             linetype = "dashed", size = 1) +
  scale_y_continuous("") +
  guides(color = "none") +
  scale_x_continuous("", limits = c(70, 120), breaks = c(94, 97)) +
  theme(axis.text.x = element_text(size = 20))
```
]

.pull-right[
#### $\large n = 15$
.small-code[
```{r echo = TRUE}
power.t.test(n = 15, 
             delta = 3, 
             sd = 5, 
             sig.level = 0.001, 
             power = NULL)
```
]
]

---
# example in `R`

.pull-left[

#### $\large \mu_1 = 94$, $\large \mu_2 = 97$

#### $\large \sigma = 5$

```{r}
delta_df <- data.frame(x = x, 
                       y1 = c(dnorm(x, 94, 5), dnorm(x, 97, 5)),
                       population = rep(c("A", "B"), each = 200))

n2 <- data.frame(x1 = c(rnorm(100, 94, 5), rnorm(100, 97, 5)),
                 y = rep(c(0, 0.001), each = 100),
                 population = rep(c("A", "B"), each = 100))
mu_df <- data.frame(population = c("A", "B"),
                    mu = c(94, 97))

ggplot(delta_df, aes(x = x, y = y1, color = population)) +
  geom_path(size = 1.75) +
  geom_point(data = n2, aes(x = x1, y = y, color = population), alpha = 0.75) +
  geom_vline(data = mu_df, aes(xintercept = mu, color = population), 
             linetype = "dashed", size = 1) +
  scale_y_continuous("") +
  guides(color = "none") +
  scale_x_continuous("", limits = c(70, 120), breaks = c(94, 97)) +
  theme(axis.text.x = element_text(size = 20))
```
]

.pull-right[
#### $\large n = 100$
.small-code[
```{r echo = TRUE}
power.t.test(n = 100, 
             delta = 3, 
             sd = 5, 
             sig.level = 0.001, 
             power = NULL)
```
]
]

---
## When should I do a power analysis?

#### Prospective is always better than retrospective!

--

**Retrospective** (conducted after experiment) 

- If you failed to reject the null, then your power was low  

- But you can't use this as an excuse!  

- Only useful as a way of planning a subsequent experiment  

--

**Prospective** (Done before the experiment)  

- Used to determine sample size or power, given $\beta$ and $\sigma$

- How can $\beta$  and/or $\sigma$ be known ahead of time?  

    + Requires prior knowledge, perhaps from a pilot study

    + Requires clear-headed thinking about what consitutes a biologically signiffcant difference

---
## What level of power should I aim for?

<br/>

#### We want power to be as close to 1 as possible

<br/>

--

#### Sometimes it may be prohibitively expensive to obtain a sample size large enough to achieve power close to 1

<br/>

--

#### In practice, we are usually satisfied with power > 0.8


---
# summary

- Power analysis let's you determine the necessary sample size (or power) for testing an effect size of interest

--

- Power is influenced by the magnitude of the effect, the standard deviation of the population, the Type I error rate, and the sample size

--

- Retrospective power analysis isn't useful unless you are planning a subsequent experiment
 
--

- `R` has several functions for conducting power analysis, but only for simple tests

--

- More complicated power analysis can be performed using simulation (not covered in this course)

---
# looking ahead

<br/>


### **Next time**: Linear models part 2: categorical predictor >2 levels

<br/>

### **Reading**: [Fieberg chp. 3.7](https://statistics4ecologists-v1.netlify.app/matrixreg#ancova)
