---
format:
  revealjs:
    highlight-style: github
    incremental: false
    theme: FANR6750.scss
    reference-location: document
knitr:
  opts_chunk:
    dev: png
    dev.args: { bg: "transparent" }
from: markdown+emoji
---

```{r setup}
#| echo: false
#| include: false
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo = FALSE, fig.align = 'center', warning = FALSE,
                      message = FALSE, fig.retina = 2)
source(here::here("R/zzz.R"))
library(FANR6750)
library(dplyr)
```

## LECTURE 6: evaluating linear model assumptions {.theme-title .center}

::: footer
FANR 6750  
Fall 2026
:::

## Outline {.theme-inverse .smaller}

<br/>

### 1) Review of linear model assumptions

<br/>  

::: {.fragment}
### 2) Evaulating assumptions

<br/> 
:::

::: {.fragment}
### 3) Robustness to assumption violations

<br/> 
:::

## Assumptions {.smaller}

[EVERY]{.highlight} model has assumptions

:::{.incremental}

- Assumptions are necessary to simplify real world to workable model


- If your data violate the assumptions of your model, inferences *may* be invalid


- [Always]{.assumptions} know (and test) the assumptions of your model
:::


## Linear model assumptions {.smaller}

</br>

$$\large y_i = \beta_0 + \beta_1 x_i + \epsilon_i$$

$$\large \epsilon_i \sim normal(0, \sigma)$$

:::{.incremental}

1) [Linearity]{.highlight}: The relationship between $x$ and $y$ is linear


2) [Normality]{.highlight}: The residuals are normally distributed


3) [Homogeneity]{.highlight}: The residuals have a constant variance at every level of $x$

4) [Independence]{.highlight}: The residuals are independent (i.e., uncorrelated with each other)

:::

## Evaluating assumptions {.smaller}

Evaluating whether your data violate the linear model assumptions relies heavily on [residuals]{.highlight} 

:::{.incremental}
- Residuals are the difference between the observed and predicted response value of each observation

    + $r_i = (y_i - \hat{y}_i)$

- Plotting the residuals vs the predicted values provides a visual assessment of the linearity assumptions

- Plotting the residuals vs the predicted values or $X$ provides a visual assessment of the variance assumption

- Plotting a Q-Q plot or histogram of the residuals provides visual assessments of the normality assumption
:::


## Evaluating assumptions {.smaller}

Fortunately, multiple `R` packages include functions for making diagnostic plots from a fitted `lm` object

```{r echo = TRUE, fig.height=5, fig.width=8}
library(performance)
data("biomassdata")
fm1 <- lm(biomass ~ elevation + rainfall, data = biomassdata)
check_model(fm1, check = c("linearity", "homogeneity", "qq", "normality"))
```


## Linearity {.smaller}

Deviations from linearity will appear as patterns in the residual plot

```{r fig.height=5, fig.width=6}
fm1_df <- data.frame(Residuals = fm1$residuals,
                     Fitted = fm1$fitted.values)
ggplot(fm1_df, aes(x = Fitted, y = Residuals)) +
  geom_point() +
  geom_hline(yintercept = 0, color = "grey40", linetype = "dashed") +
  stat_smooth()
```

```{r fig.height=5, fig.width=6}
x <- runif(50, 0, 10)
ey <- x^1.5
y <- rnorm(50, ey, 1)
fm2 <- lm(y ~ x)
fm2_df <- data.frame(Residuals = fm2$residuals,
                     Fitted = fm2$fitted.values)
ggplot(fm2_df, aes(x = Fitted, y = Residuals)) +
  geom_point() +
  geom_hline(yintercept = 0, color = "grey40", linetype = "dashed") +
  stat_smooth()
```


## Homogeneity of variance {.smaller}

Heteroscedasticity will appear as patterns (often funnel-shaped) in the residual plot

```{r fig.height=5, fig.width=6}
fm1_df2 <- data.frame(Residuals = fm1$residuals,
                     Fitted = fm1$fitted.values)
ggplot(fm1_df2, aes(x = Fitted, y = Residuals)) +
  geom_point() +
  geom_hline(yintercept = 0, color = "grey40", linetype = "dashed") +
  stat_smooth()
```

```{r fig.height=5, fig.width=6}
x <- runif(50, 0, 10)
ey <- 2 + 1.5 * x
sig <- 0.25 * 0.75 * x
y <- rnorm(50, ey, sig)
fm3 <- lm(y ~ x)
fm3_df <- data.frame(Residuals = fm3$residuals,
                     Fitted = fm3$fitted.values)
ggplot(fm3_df, aes(x = Fitted, y = Residuals)) +
  geom_point() +
  geom_hline(yintercept = 0, color = "grey40", linetype = "dashed") +
  stat_smooth()
```



## Homogeneity of variance {.smaller}

Heteroscedasticity can also be assessed by plotting the square root of residuals. Violations will appear as increasing or decreasing trends

```{r fig.height=5, fig.width=6}
fm1_r <- sqrt(abs(fm1$residuals/sd(fm1$residuals)))
fm1_df2 <- data.frame(Residuals = fm1_r,
                     Fitted = fm1$fitted.values)
ggplot(fm1_df2, aes(x = Fitted, y = Residuals)) +
  geom_point() +
  geom_hline(yintercept = 0, color = "grey40", linetype = "dashed") +
  stat_smooth() +
  scale_y_continuous("sqrt(|stn. r|)")
```

```{r fig.height=5, fig.width=6}
fm3_r <- sqrt(abs(fm3$residuals/sd(fm3$residuals)))
fm3_df2 <- data.frame(Residuals = fm3_r,
                     Fitted = fm3$fitted.values)
ggplot(fm3_df2, aes(x = Fitted, y = Residuals)) +
  geom_point() +
  geom_hline(yintercept = 0, color = "grey40", linetype = "dashed") +
  stat_smooth() +
  scale_y_continuous("sqrt(|stn. r|)")
```


## Normality {.smaller}

Normality is usually assessed using *quantile-quantile* (Q-Q) plots, which plots the observed quantiles of the residuals vs. the expected quantiles of normally distributed data 

```{r fig.height=5, fig.width=6}
ggplot(biomassdata, aes(sample = biomass)) +
  stat_qq() +
  stat_qq_line() +
  scale_y_continuous("Sample quantiles") +
  scale_x_continuous("Standard normal quantiles")
```

```{r fig.height=5, fig.width=6}
q_df <- data.frame(x = runif(60),
                   eps = rlnorm(60)) %>%
  mutate(ey = 3.2 + 1.2 * x,
         y = ey + eps)

ggplot(q_df, aes(sample = y)) +
  stat_qq() +
  stat_qq_line() +
  scale_y_continuous("Sample quantiles") +
  scale_x_continuous("Standard normal quantiles")
```


Large deviations from the line indicate deviations from normality

## Evaluating models with factors {.smaller}

Because all of the models we have seen so far this semester are linear models, they all share the same assumptions


Testing assumptions of models with factors (*t*-test, ANOVA) is therefore no different than testing assumptions of models with continuous predictors

```{r echo = TRUE, fig.height=3, fig.width=6}
data("tunadata")
fm2 <- lm(growth ~ status, data = tunadata)
```

```{r fig.height = 3.5, fig.width=7.5}
qq <- ggplot(tunadata, aes(sample = growth)) +
  stat_qq() +
  stat_qq_line() +
  scale_y_continuous("Sample quantiles") +
  scale_x_continuous("Standard normal quantiles")

fm2_df <- data.frame(Residuals = sqrt(abs(fm2$residuals/sd(fm2$residuals))),
                     Fitted = fm2$fitted.values,
                     X = tunadata$status)
h <- ggplot(fm2_df, aes(x = X, y = Residuals)) +
  geom_boxplot() +
  stat_smooth(method = "lm") +
  scale_x_discrete("Status") +
  scale_y_continuous("sqrt(|stn. r|)")

cowplot::plot_grid(h, qq, nrow = 1)
```


## Null hypothesis testing {.smaller}

It is also possible to "test" whether assumptions are met:

```{r echo = TRUE, fig.height=3, fig.width=6}
check_normality(fm1)
check_heteroscedasticity(fm1)
```

where the null hypothesis is that the data don't violate the assumption being test (so $p>0.05$ is good)


However, remember that the assumptions are about the population, not the sample

- samples can deviate from assumptions, even if they are met for the population

- deviations are more likely for small sample sizes

- there is always some judgement necessary when testing assumptions

We will explore some of these tests in lab


## Independence {.smaller}

Non-independence of residuals is more difficult to test but is often related to un-modeled structure in the data

- points close in space/time tend to be more alike than distance points

- observations within "groups" (e.g., study plots, growth chamber, lakes) tend to be more alike than individuals from other groups


Often, non-independence can be judged based on knowledge of the sampling design/system

- sometimes can be seen as "clustering" in residual plots

```{r fig.width = 4, fig.height=2.5}
x <- rnorm(50)
group <- sample(1:5, 50, replace = TRUE)
nu <- c(-5, -2, 0, 2, 5)
ey <- 5  + nu[group] + (1.2) * x
y <- ey + rnorm(50, 0, 0.5)

fm <- lm(y ~ x)

df <- data.frame(x = x, y = y, r = fm$residuals, 
                 e = fm$fitted.values, Lake = as.factor(group))
ggplot(df, aes(x = e, y = r, color = Lake)) + 
  geom_point() +
  scale_x_continuous("Expected value") +
  scale_y_continuous("Residuals") +
  theme(legend.title = element_text())
```


## Independence {.smaller}

```{r fig.width = 6, fig.height=4}
ggplot(df, aes(x = e, y = r, color = Lake)) + 
  geom_point() +
  scale_x_continuous("Expected value") +
  scale_y_continuous("Residuals") +
  theme(legend.title = element_text())
```

Sometimes, non-independence can be remedied by including relevant covariates in the model

- e.g. `lm(y ~ x + Lake)`

- we'll learn more about this and related approaches in a few weeks

## What happens when assumptions are violated? {.smaller}

One consequences of sampling is that our samples will likely depart from assumptions to some degree

So rather than asking:

> "Do my data violate the assumptions of my model?"

It can be useful to ask:

> "Do my data violate the assumptions of my model *enough to change my conclusions*?"

Fortunately, linear models are relatively robust to minor (or even moderate) departures from the assumptions

## What happens when assumptions are violated? {.smaller}

General rules of thumb:

- Minor assumption violations will not generally bias estimates of $\hat{\beta}_0$, $\hat{\beta}_1$, etc.

- For large sample sizes, Central Limit Theorem often ensures that residuals of many types of data are normally distributed (or at least pretty close)

- Standard errors (and therefore confidence intervals, p-values, etc) *will* be biased when assumptions are violated

- In general, violating assumptions results in standard errors that are too small. What does this do to p-values?

## Looking ahead

<br/>

[Next time]{.highlight}: Principles of Causal Inference

<br/>

[Reading]{.highlight}: [Causal Inference in R, chp 3](https://www.r-causal.org/chapters/03-po-counterfactuals)
