---
title: "LECTURE 20: zero-inflated data"
subtitle: "FANR 6750 (Experimental design)"
author: "<br/><br/><br/>Fall 2023"
output:
  xaringan::moon_reader:
    css: ["default", "FANR6750.css", "FANR6750-fonts.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, echo = FALSE, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo = FALSE, fig.align = 'center', warning=FALSE, message=FALSE, fig.retina = 2)
source(here::here("R/zzz.R"))
library(emo)
library(FANR6750)
library(dplyr)
library(kableExtra)
# library(gganimate)
```

# outline 


1) Motivation

<br/>

--

2) Negative binomial regression

<br/>

--

3) Hurdle models

<br/>

--

4) Zero-inflated mixture models


---
# poisson regression

Poisson regression is a specific type of GLM in which the response variable follows a Poisson distribution and the link function is the log  

--

Poisson regression is used to model a positive, integer response variable as a function of explanatory variables

--

#### Examples:

- Is the abundance of a focal species related to habitat type?  

- Is the number of foraging bouts a function of predator density?  

- Is flock size influenced by food availability?  

---
# poisson distribution

<br/>

```{r fig.width=8, fig.height=4}
x <- seq(0, 12, by = 1)
df <- data.frame(y = c(dpois(x, lambda = 0.1), dpois(x, lambda = 1), dpois(x, lambda = 3)),
                 x = x,
                 lambda = factor(rep(c("lambda = 0.1", "lambda = 1", "lambda = 3"), each = length(x)),
                                 levels = c("lambda = 0.1", "lambda = 1", "lambda = 3")))
ggplot(df, aes(x = x, y = y, fill = lambda)) +
  geom_col(alpha = 0.75, color = "grey", linewidth = 0.1, position = "dodge") +
  scale_y_continuous("Probability density") +
  scale_x_continuous("y") +
  scale_fill_discrete(labels = c(expression(paste(lambda, "= 0.1")), 
                                 expression(paste(lambda, "= 1")), 
                                 expression(paste(lambda, "= 3"))))
```

---
# example

Imagine we are interested in restoring oyster beds along the Georgia coast. We "seed" experimental plots with larvae and then record the number of oysters that establish within a randomly-selected quadrat within each plot. We also record the substrate (rock vs sand), water temperature, water depth, salinity, and water clarity of each plot

.pull-left[
.vsmall-code[
```{r echo = TRUE}
data("oysterdata")
head(oysterdata)
```
]
]

.pull-right[
<br/>
```{r fig.width=4, fig.height=3}
ggplot(oysterdata, aes(x = oyster)) +
  geom_histogram() +
  scale_y_continuous("Frequency") +
  scale_x_continuous("Number of oysters")
```
]

---
# zero inflation

Zero inflation occurs when there are more zeros in the response variable $y$ than expected under the assumed probability distribution (usually the Poisson distribution)

--

.pull-left[
.small-code[
```{r echo = TRUE}
fit.pr <- glm(oyster ~ .,
              data = oysterdata,
              family = "poisson")

```
]
]

.pull-right[
.vvsmall-code[
```{r echo = TRUE}
summary(fit.pr)
```
]
]


---
# zero inflation

For models with with the `glm()` function, we can test for zero inflation:

.vsmall-code[
```{r echo = TRUE, message=TRUE}
performance::check_zeroinflation(fit.pr)
```
]

--

If we suspect zero inflation, we have several options:

- Negative binomial regression

- Hurdle models

- Zero-inflated mixture model

---
class: inverse, center, middle

# negative binomial regression

---
# negative binomial distribution

The negative binomial distribution is a *discrete* probability distribution (only integers) with support (i.e., possible values) from 0 to $+\infty$

--

The negative binomial distribution has two parameters: $\mu$ and $\theta$

--

- The mean of the negative binomial distribution is $\mu$

- The variance is $\mu + \frac{\mu^2}{\theta}$

- As $\theta \to \infty$, the negative binomial distribution converges to the Poisson

--

In `R`:

.vsmall-code[
```{r echo = TRUE}
x <- rnbinom(n = 1000, mu = 1, size = 1) # theta = size
mean(x)
var(x)
```
]

---
# negative binomial distribution

Because the variance is larger than the mean, the negative binomial distribution is often used to model over-dispersed data, including zero inflation

--

.left-nudge[
.small-code[
```{r echo = TRUE}
library(MASS)
fit.nb <- glm.nb(oyster ~ ., 
                data = oysterdata)
```
]
]

.right-nudge[
.vvsmall-code[
```{r echo = TRUE}
summary(fit.nb)
```
]
]

---
# negative binomial regression

Did the negative binomial model address the zero inflation?

--

.small-code[
```{r echo = TRUE, message=TRUE}
performance::check_zeroinflation(fit.nb)
```
]

--

Although the `check_zeroinflation()` indicates some remaining zero inflation, the negative binomial model seems to fit relatively well 

- additional tests are possible and probably warranted. See Fieberg chp. 17.3 for additional detail

--

Sometimes, rather than address zero inflation using a different distribution, we may suspect that (some of the) zero's are the result of a *different process* than the one generating the counts

- What are some examples?

- In these cases, models that treat zeros as a mixture of different distributions may be warranted

---
class: inverse, center, middle

# hurdle models

---
# hurdle models

Hurdle models are actually composed of two separate models, one estimating the probability that an observation is 0 and another modeling the non-zero counts

--

- What distribution do you think is used for the zero/non-zero part of the model?

--

    + Assume $z_i = 0$ if $y_i = 0$, otherwise $z_i = 1$


$$z_i \sim Bernouli(p_i)$$

$$logit(p_i) = \beta_0 + \beta_1 x_{i,1} + \beta_2 x_{i,2}...$$
--

- What distribution do you think is used for the non-zero observations?

--

    + zero-truncated Poisson/Negative binomial

---
# zero-truncated distributions

<br/>

```{r fig.width=8, fig.height=5}
dtpois<-function(x, lambda){
    0 + I(x>0)*dpois(x, lambda)/(1-dpois(0, lambda))
}
x<-seq(0,10, by=1)
ztp <- data.frame(x = x, 
                  y = c(dpois(x, lambda = 2), dtpois(x, lambda = 2)),
                  dist = rep(c("Poisson", "Zero-truncated Poisson"), each = length(x)))

ggplot(ztp, aes(x = x, y = y, fill = dist)) +
  geom_col(alpha = 0.75, color = "grey", linewidth = 0.1, position = "dodge")+
  # facet_wrap(~dist, nrow = 1) +
  labs(title = expression(paste(lambda, " = 2"))) +
  scale_y_continuous("Density") +
  scale_x_continuous("y", breaks = c(0, 2, 4, 6, 8, 10))
```

---
# hurdle models

.left-nudge[
.vsmall-code[
```{r echo = TRUE}
library(pscl)
fit.hrp <- hurdle(oyster ~ ., 
                 data = oysterdata)
```
]
]

.right-nudge[
.vvsmall-code[
```{r echo = TRUE}
summary(fit.hrp)
```
]
]

---
# hurdle models

.left-nudge[
.vsmall-code[
```{r echo = TRUE}
fit.hrnb <- hurdle(oyster ~ ., 
                 data = oysterdata,
                 dist = "negbin")
```
]
]

.right-nudge[
.vvsmall-code[
```{r echo = TRUE}
summary(fit.hrnb)
```
]
]


---
# hurdle models


Sub-models can contain the same or different covariates

--

Parameters for sub-models are interpreted in the same way as for logistic/Poisson regression

--

$\lambda_i$ is the expected value (i.e., mean) of the non-zero observations, not the overall mean of all observations (which is $\frac{1-p_i}{1-e^{\lambda_i}}\lambda_i$)

--

Models can be compared using AIC

  - note that `performance::check_zeroinflation()` does not work for `hurdle` model objects

--

Generally, Hurdle models are most appropriate when the count-based process cannot generate zeros (i.e., zeros are determined by a separate process than counts)

- Examples?

---
class: inverse, center, middle

# zero-inflated mixture models

---
# zero-inflated mixture models

Sometimes, zero observations can occur due to both the binomial process and the count process

- Habitat may be suitable or unsuitable, but abundance may be zero at suitable locations (especially if density is low)

- Individuals may be breeders or non-breeders, but some breeders will still fail to produce offspring (e.g., nest predation)

--

In these cases, the zero-truncated distribution used by the Hurdle model is not appropriate

--

Zero-inflated mixture models allow zero observations to arise from both a logistic/binomial process and a count model

- Both Poisson and negative binomial count models are possible


---
# zero-inflated mixture models

.left-nudge[
.vsmall-code[
```{r echo = TRUE}
fit.zip <- zeroinfl(oyster ~ .|., 
                 data = oysterdata)
```
]
]

.right-nudge[
.vvsmall-code[
```{r echo = TRUE}
summary(fit.zip)
```
]
]

---
# zero-inflated mixture models

.left-nudge[
.vsmall-code[
```{r echo = TRUE}
fit.zinb <- zeroinfl(oyster ~ .|., 
                 data = oysterdata,
                 dist = "negbin")
```
]
]

.right-nudge[
.vvsmall-code[
```{r echo = TRUE}
summary(fit.zinb)
```
]
]

---
# model comparison

Generally, alternative zero-inflated models are compared using AIC

```{r}
aictab <- data.frame(Model = c("Poisson GLM", "NegBin GLM", "Poisson Hurdle",
                                "NegBig Hurdle", "Zero-inflated Poisson", 
                                "Zero-inflated NegBig"))
aictab <- dplyr::bind_cols(aictab, AIC(fit.pr, fit.nb, fit.hrp, fit.hrnb, fit.zip, fit.zinb))
aictab$dAIC <- aictab$AIC - min(aictab$AIC)
aictab <- dplyr::arrange(aictab, dAIC)

kable(aictab, col.names = c("Model", "k", "AIC", "delta AIC"), 
      row.names = F) 
```

--

Do AIC scores tell us if the models are "good"?

--

- NO!

- Goodness-of-fit tests are also important for zero-inflated model. See [Fieberg 17.7](https://fw8051statistics4ecologists.netlify.app/zimodels#goodness-of-fit) for examples

---
# looking ahead

<br/>

### **Next time**: Exam 3 review



