---
title: "Lab 3: Introduction to Linear Models"
subtitle: "FANR 6750: Experimental Methods in Forestry and Natural Resources Research"
author: "Fall 2023"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{lab03_linear_models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, fig.retina = 2,
                      fig.align = "c",
  collapse = TRUE,
  comment = "#>"
)
library(dplyr)
library(knitr)
library(FANR6750)
library(ggplot2)
#source(here::here("R/zzz.R"))
```

## Lab 2

- Introduction to R Markdown

- Creating R Markdown files

- Creating code chunks

- Using RStudio Projects

- Organizing directories and subdirectories


## Today's topics

- Introduction to linear models

- Simple Linear Regression

  + Parameter interpretation
  
  + Residuals
  
  + Assumptions of linear models

- Linear models in R

  + A brief aside about R graphics

  + Parameter interpretation in R

<br>
<br>

# What is a Linear Model?

Formally, a linear model is defined as **a model that specifies a linear relationship between a response variable and a number of predictor variables.**

<br>

Below is an example of a linear model. We will call this the "General Linear Model" because it does not represent a specific model but rather represents all linear models in general.

$$
\overbrace{{y_i}}^\text{response} = \underbrace{\color{red}{\beta_0} + \color{red}{\beta_1}{x_{i1}} + \color{red}{\beta_2}{x_{i2}} + ... + \color{red}{\beta_p}{x_{ip}}}_\text{Deterministic} + \overbrace{\color{red}{\epsilon_i}}^\text{stochastic}
$$

where:

$y$ = the response

$\beta$ = the effect that a particular predictor has on the response

$x$ = a predictor

$\epsilon$ = residuals

<br>

Note: Some of the values are highlighted in $\color{red}{red}$. This is showing the fact that these values are estimated **from** the model whereas the other values are provided **to** the model in the form of data.

<br>

Note: The deterministic part is the part that we can account for given the data that we have. The stochastic portion is the extra bit (randomness) for which our model does not account.

<br>

Also keep in mind that the predictor variables ($x_i$) may take many forms. What are some examples of the types below?

- Continuous
- Binary
- Factors

<br>
<br>

# Simple Linear Regression

Let's look at an example of a linear model with only one continuous predictor. In this dataset, the yearly rainfall (mm) is being used to predict the average number of acorns produced by White Oaks (*Quercus alba*) in north Georgia.

$$
y_i = \beta_0 + \beta_1x_i + \epsilon_i \ \ \ \ \ \text{where} \ \ \ \ \ \epsilon_i \sim \text{N}(0,\sigma)
$$
Here we have an example where the response variable (y) is being predicted using only one predictor variable (x). 

```{r, include= F}
library(car)
library(MASS)
library(lmtest)
library(tseries)
library(ggfortify)

set.seed(1234)

##Linear Regression
#Generate the independent variable and the error
rainfall=round(rnorm(100,50,8),1)
error=rnorm(100,0,45.33)
#Generate the dependent variable
acorns=round(10000+(1.663*rainfall)+error,0)
#create the model
m1=lm(rainfall~acorns)
raindata <- as.data.frame(cbind(rainfall, acorns))
write.csv(raindata, 'raindata.csv', row.names= F)
```

```{r, echo= F, fig.width= 8, fig.height= 4, fig.fullwidth= T}
ggplot(data= raindata, aes(x= rainfall, y= acorns)) +
  geom_point(color= 'blue') +
  geom_smooth(method= 'lm', se= F)
```

## Parameter Interpretation

Above, we have written a linear model which predicts the number of acorns on an average White Oak tree using the yearly rainfall. Before moving forward, let's talk about how we would interpret the $\beta_0$ and $\beta_1$ in our model. 

<br>

$\beta_0$: Intercept; the expected value of $y_i$ when all predictors = 0

$\beta_1$: Slope: the expected change in $y_i$ resulting from one unit change in the associated predictor variable $x_n$

<br>
<br>

A few questions to think about related to these:

1) What is the biological meaning of the intercept? Does it have a biological meaning?

2) What are the units of the predictor variables (e.g., mm, $\mu L$, % canopy cover)?

<br>

## Residuals

Keep in mind that in any linear model that we create, we will not be able to perfectly predict a response given a set of predictor variables. This is especially true in ecological systems where data are often quite messy. As a result, we can see above that even though we have fitted a prediction line through the dataset, it overestimates the number of acorns at some values of rainfall and underestimates at others. 

<br>

These differences between the observed values of the response ($y_i$) and the predicted values of the response from our model ($E[y_i]$) are known as **Residuals**. These residuals represent the variation in $y_i$ which is not explained by any $x_i$ in our model.

<br>

Later in the lab, when we attempt to estimate $\beta_0$ and $\beta_1$ above, we will do this by minimizing the size of the residuals (e.g., ordinary least squares).

<br>

## Assumptions of linear models

When we are using `R` to help with data analysis, keep in mind that, for the most part, `R` is happy to do anything that you tell it to do. In reality though, every linear model that we will use this semester will come with a set of assumptions. Before using any model for data analysis, you will need to consider the assumptions of that model and decide whether or not it is reasonable to make those assumptions. If it is, that model may be appropriate for that analysis; if it is not, you will need to consider other options.

All linear models have at least 6 explicit assumptions. In this course, we will primarily consider the first 4 of these assumptions, but it is good to know that the others (and usually more) exist. As we progress through the course, we will see how we can test these assumptions and what our options are when the assumptions are not met.

As we go through these assumptions try to find in our linear model where these assumptions show up (some are more subtle than others).


1) **Linearity:** We assume that the relationship between x and y is linear

<br>

Note: This assumption **DOES NOT** mean that we need to be able to fit a straight line through the data. Often, we will actually used a curved line (in the cases of quadratic relationships). Rather, this assumption means that the **parameters** themselves enter the model linearly. This means that for one unit change in a predictor, there is **ALWAYS** the same constant change in the response.

Let's see some examples:

$$
y_i = \beta_0 + \beta_1x_1 + \epsilon_i
$$

$$
y_i = \beta_0 + \beta_1x_1 + \beta_2x_1^2 + \epsilon_i
$$

$$
y_i = \beta_0 + \beta_1x_1 + \beta_2x_2x_3 + \epsilon_i
$$

$$
y_i = \beta_0 + x_1^{\beta_1} + \beta_2x_1^2 + \epsilon_i
$$
<br>

2) **Normality:** We assume that the residuals are normally distributed

This assumption will be important as we progress through the semester and perform hypothesis testing to assess the significance of predictor variables. While normality of residuals is not required, it suggests that the model is doing a good job of predicting the response (i.e. most residuals are close to zero and larger errors are much less likely).

<br>
<br>

3) **Homoscedasticity:** We assume that the residuals have a constant variance across all levels or values of x (if not, we refer to that as heterscedasticity)

Variance of the residuals can best be seen by creating a plot of the model residuals across the range of a predictor.

```{r, echo= F, fig.width= 8, fig.height= 4, fig.fullwidth= T}
set.seed(4)
n <- 1000
x <- seq(1,100,length.out = n)
residuals <- 2 + 2*x + rnorm(n, sd = 2*abs(x)^1.1)
d <- data.frame(residuals, x)
ggplot(d, aes(x, residuals)) + geom_point()
```
<br>
<br>

When might this assumption be violated?

- In factors
- In time series
- In cross sectional studies

<br>
<br>

4) **Independence:** We assume that the residuals are independent of one another and independent of predictors.

This assumption means that we cannot predict the size of a residual from another residual (autocorrelation) or from the value of a predictor variable (endogeneity).

When might this assumption be violated?

<br>
<br>

5) **Unbiased estimator:** We assume that the residuals are centered on zero. 

If the residuals are centered on zero, then our estimate of the parameter is doing a good job of approximating the true value of the parameter. If the residuals are not centered on zero, then the model is either consistenly under or over estimating the parameter values (a biased estimator).

```{r, echo= F, fig.width= 8, fig.height= 4, fig.fullwidth= T}
set.seed(4)
residuals <- rnorm(10000, 0.7, 1.4)
hist(residuals, breaks= 9, main= 'Histogram of Residuals', xlim= c(-10,10))
```

<br>
<br>

6) **Independent predictors:** We assume that when using multiple predictor variables, they are independent from one another.

If one explanatory variable is able to be predicted using others (the information provided by the variable is redundant) then we refer to that as multicollinearity. 

<br>
<br>

What are some examples?

<br>
<br>

- Predicting nest survival using both Fahrenheit and Celsius

<br>

- Predicting presence of birds using % land use categories

<br>

- Predicting movement of fish using dissolved oxygen, salinity, and conductivity

<br>
<br>

# Linear models in R

Above, we have seen an example of a linear model with only one continuous predictor variable. Now we will see how we might create this model in `R`.

## Exercise 1

1) Open your 'FANR6750' Rstudio project (if you have one)

2) Create a new 'R' script and save it to the directory where you store your lab activities. Name it something like 'lab03-SLR.R'

3) Load the 'FANR6750' package and the 'raindata' object

```{r}
library(FANR6750)
data('raindata')
```

Before we do anything else, lets examine our dataset and plot the rainfall-acorn relationship.

## Graphics

`R` has very powerful graphing capabilities that make it possible to create data visualizations for reports or publications. As with most tasks in `R`, there are many ways to create graphs and you will find that people have very strong feelings about the best approach. 

The debate over graphics in `R` usually boils down to using the built-in graphing functions ("base graphics") vs the `ggplot2` package. There are some advantages to both approaches, but for the most part the newer generations of `R` users prefer `ggplot2`. Therefore most of the sample code provided in lab will reflect that preference. However, **I don't care how you make your plots as long as they effectively display the information you are trying to convey**. If you prefer base graphics, by all means use base graphics.^[Keep in mind that although we will be using ggplot2, it is a good idea to have some understanding of how base `R` graphics work. Most packages in `R`, when they produce some sort of output in the form of a plot, will still create that plot using base `R` graphics rather than ggplot. In order to alter those graphics without having to start from scratch, it will be useful to know a little about base `R` graphics.]  

### Brief introduction to `ggplot2`

Because the code I provide will use `ggplot2`, it is worth briefly learning/reviewing how this package approaches data visualization. 

The power and flexibility of `ggplot2` come from it's consistent structure. Although a bit confusing at first, once you get the hang of it, the structure actually makes it quite easy to create highly customized visualizations. All plots created using `ggplot2` use the same underlying structure:

$$\underbrace{ggplot}_{initiate\; plot}(\underbrace{data = df}_{data\;frame},\; \underbrace{aes(x =\; , y = \;)}_{plot\; attributes}) + \underbrace{geom\_line()}_{geometry}$$

The `ggplot()` function initiates a new plot. In this function, you tell `ggplot2` what data frame you will be using for the plot (`ggplot` only accepts data frames as input) and you tell it how to map attributes of the data to the visual properties of the figures. Attributes are mapped inside the `aes()` argument. Attributes usually include location (`x-axis` and `y-axis` placement), color, size, shape, line type, and many others. In general, **each attribute will be mapped to one column of your data frame**. 

The `ggplot()` function simply initiates a graph - if you run just that portion of the code you will get a blank graph. We can see that by creating a new plot showing the relationship between `rainfall` (the x-axis of the plot) and `acorns` (the y-axis): 



```{r fig.width=5, fig.height=4, fig.align='center'}
library(ggplot2)
ggplot(data = raindata, aes(x = rainfall, y = acorns))
```

You can see that `ggplot` created a figure with the correct axes and labels. But no data. That's because we didn't tell `ggplot` what type of *geometry* to use to represent the data. Geometry refers to the type of geometric object(s) we want to use to display the data. Common geometries include points (e.g., scatter plot), lines (e.g., time series), and bars (e.g., histograms). There are many others. Once we add a geometry, we can see the data:

```{r fig.width=5, fig.height=4, fig.align='center'}
ggplot(data = raindata, aes(x = rainfall, y = acorns)) +
  geom_point()
```

# Back to Simple Linear Regression

Now that we have plotted the data, we can see that there appears to be a linear relationship between the rainfall in a particular year and the acorn production of White Oaks. Next, we will perform Simple Linear Regression in `R` to estimate the intercept and slope of the regression line. 

```{r}
mod1 <- lm(acorns~ rainfall, data= raindata)

# It is often helpful to adjust the way R displays numbers to avoid having to 
# interpret scientific notation. One way we can do that is using the options function.

options(scipen= 100)

summary(mod1)
```

## Parameter interpretation in R

One of the most important parts of using `R` for data analysis will be learning to interpret the results provided as output.

How should we interpret these model parameters? What do they mean in the context of the statistical model and in the context of the biological data?

```{r fig.width= 8, fig.height= 4, fig.fullwidth= T}
# Now let's plot our model predictions
mod1_fit <- predict(mod1)
ggplot(data = raindata, aes(x = rainfall, y = acorns)) +
  geom_point() +
  geom_line(aes(x = rainfall, y= mod1_fit))
```

<br>
<br>
<br>

# Assignment

In this data example, a researcher is interested in using the average density of feral cats (cats/square mile) in 100 city parks across the US to predict the number of lizards/square mile (prey) at those parks.

Create an R Markdown file to do the following:

1. Create an `R` chunk to load the prey data using:

```{r, include= F}
set.seed(1234)

##Linear Regression
#Generate the independent variable and the error
predators=round(rnorm(100,17,2.2),1)
prey=rnorm(100,800, 366.8)
year=rep(2006:2015, each= 10)
error=rnorm(100, 49.7, 15.578)
#Generate the dependent variable
prey=round(80.5944-(4.733*predators)+error, 1)
#create the model
m1=lm(prey~ predators)
preydata <- as.data.frame(cbind(year, predators, prey))
plot(prey~ predators, data= preydata)
abline(m1)
write.csv(preydata, 'preydata.csv', row.names= F)
```

```{r eval = FALSE}
library(FANR6750)
data("preydata")
```



2. Make a figure that plots the raw data. Keep in mind which axis you will want to plot the predictor variable vs the response variable.

3. Create a content header called "Linear Model" and under this header, use R Markdown's equation editor to write the linear model you will use to help answer this researcher's question. Remember to include all parts of the model including how you would specify the residual structure. Below the model, make sure to define what every term in the model means.

4. Create a header called "Linear Models in R". Use the `lm()` function to translate your linear model into `R` and look at the summary of results.

5. Interpret the results of your model. Explain what each parameter means both in terms of the model as well as biologically.

6. Make a new plot which shows both the raw data as well as the fitted line from your model.

7. Add a section at the end of your file called "Discussion" and address the following questions.

  a. What would happen if we tried to use this model to make predictions outside the range of our data? Would it still be a useful model? Why or why not?
  
  b. How many lizards would there be if the density of cats = 80? Explain what is happening here and how you might resolve this issue?
  
  c. Below are two figures. The first shows the residuals of the model plotted against the predicted (fitted) values along with a horizontal line showing the mean of the residuals. The second plot is a histogram of the residuals. Which assumptions of linear models do these plots address? Based on what you see, do these assumptions appear to be appropriate?
  
```{r, echo= F, fig.show= 'hold'}
plot(m1$fitted.values, m1$residuals)
abline(h=mean(m1$residuals))
hist(m1$residuals)
```


A few things to remember when creating the document:

- Be sure the output type is set to: `output: html_document`

- Be sure to set `echo = TRUE` in all `R` chunks so that all code and output is visible in the knitted document

- Regularly knit the document as you work to check for errors

- See the R Markdown reference sheet for help with creating `R` chunks, equations, tables, etc.
