---
title: "Lab 13: Generalized linear models: Poisson Regression"
subtitle: "FANR 6750: Experimental Methods in Forestry and Natural Resources Research"
author: "Fall 2023"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{lab13_glm_pr}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  fig.align = 'center', fig.height = 4, fig.width = 6,
  fig.retina = 2,
  warning = FALSE, message = FALSE, cache= T,
  collapse = TRUE,
  comment = "#>"
)
source(here::here("R/zzz.R"))
library(kableExtra)
```

## Lab 12

-   Logistic Regression

    -   Link functions

    -   Using glm()

    -   Creating plots

## Today's topics

-   Poisson Regression

    - Link functions
    
    - Creating plots
    
    - Poisson vs Negative binomial

<br> <br>

# Poisson regression

This week we will see another situation where using a linear model will
not be the most appropriate approach. Here, we are interested in using
count data as our response variable and estimating counts as a function
of some predictors.

<br>

Let's take a look at a data example to demonstrate this concept. The data are modified from [Ver Hoef and Boveng 2007](https://digitalcommons.unl.edu/cgi/viewcontent.cgi?article=1141&context=usdeptcommercepub) and include 400 observed counts of harbor seals from aerial surveys conducted in coastal Alaska.


```{r eval= T}
library(FANR6750)
data("harbordata")
head(harbordata)
```

First let's visualize the data.

```{r, fig.cap='**Figure 1: Counts of harbor seals at sites across a range of tide heights (relative to low tide).**'}
library(ggplot2)

ggplot() +
  geom_point(data = harbordata, aes(x = Reltolow, y = Number)) +
  scale_y_continuous("Count") +
  scale_x_continuous("Tide relative to low (m)")
```

<br>

Here we are interested in estimating the count of harbor seals during aerial surveys. Like last week, we could start by constructing a model that estimates count as a function of tide height.

```{r}
mod1 <- lm(Number~ Reltolow, data= harbordata)
summary(mod1)
```

Now let's add our model predictions to the plot from above.

```{r, fig.cap='**Figure 2: Counts of harbor seals at sites across a range of tide heights (relative to low tide). Line represents model predictions from simple linear regression (mod1)**'}
model_pred <- predict(mod1)

ggplot() +
  geom_point(data = harbordata, aes(x = Reltolow, y = Number)) +
  geom_line(aes(harbordata$Reltolow, model_pred), color= 'red') + 
  scale_y_continuous("Count") +
  scale_x_continuous("Tide relative to low (m)")
```

<br>

**What issues do you see with this plot? Does this look like a
reasonable way to model this response variable?**

Similar to last week, we are again presented with a problem. We have an
integer count response variable but using a linear model results in
unreasonable predictions (since it is not possible for the response to
take on values in between the integers or below zero). We can resolve
this in a similar way as last week. In that lab, Our data was (0,1) and
we needed to create a response variable that could take on values only
between zero and one (inclusive) so probability was a reasonable choice.
In this case, our response is all possible positive integers but we need
a response that can take on all possible positive values. In this case,
we can model the mean response ($\lambda$) rather than the count
observations themselves. We could derive an equation that estimates
$\lambda$ as a function of our predictor variables:

$$
\lambda = e^{\beta_0 + \beta_1x1 + ... + \beta_zx_z}
$$

<br> <br> Now we've run into another problem. **What is the issue with
this equation in the context of linear modeling?**

<br> <br>

Instead, we can use a link function (in this case the log link) to
create a linear relationship between the response and the predictors.

Remember from lecture that the model for the Poisson regression is:

$$\Large log(\lambda_i) = \beta_0 + \beta_1x_{i1} + \beta_2x_{i2} + · · · + \beta_px_{ip}$$
$$\Large y_i \sim Poisson(\lambda_i)$$

where $\lambda_i$ is the expected value of $y_i$

Using the harbor data, let's model the number of harbor seals detected
at each site as a function of site substrate and tide height relative to
low tide:

```{r}
fm2 <- glm(Number ~ substrate + Reltolow, family = poisson(link = "log"), data = harbordata)
summary(fm2)
```

Notice that we use the `family =` argument to tell `R` that this is a
poisson glm with a log link function.

**What do the parameter estimates from this model represent, including
the intercept? What can we say about the effects of substrate and tide
height on harbor seal counts?**

<br> <br>

As we have seen previously, we can use `predict()` to estimate counts at
different combinations of tide height and substrate. For example, how
does the predicted count change across tide heights on ice:

```{r}
predData <- data.frame(Reltolow = seq(min(harbordata$Reltolow), max(harbordata$Reltolow), length = 10000), substrate= 'ice')
```

To get confidence intervals on the $(0, \infty)$ scale, predict on the
log (link) scale and then back transform using the inverse-link
(`exp()`):

```{r, fig.cap='**Figure 3: Counts of harbor seals at sites on ice across a range of tide heights (relative to low tide). Line represents model predictions from poisson regression model and dashed lines represent approximate 95% confidence bands.**'}
pred.link <- predict(fm2, newdata = predData, se.fit = TRUE)
predData$lambda <- exp(pred.link$fit) # exp is the inverse-link function
predData$lower <- exp(pred.link$fit - 1.96 * pred.link$se.fit)
predData$upper <- exp(pred.link$fit + 1.96 * pred.link$se.fit)

ggplot() +
  geom_point(data = harbordata, aes(x = Reltolow, y = Number)) +
  geom_path(data = predData, aes(x = Reltolow, y = lambda), color= 'red') +
  geom_ribbon(data = predData, aes(x = Reltolow, ymin = lower, ymax = upper),
              fill = NA, color = "red", linetype= 'dashed') +
  scale_x_continuous("Tide relative to low (m)") +
  scale_y_continuous("Expected count")
```

<br>
<br>

Although we have completed the analysis, this still may not be the most appropriate way to consider this dataset. **What are some issues that you see either from the plot or from the model output related to concepts discussed in lecture?**   

**Why does the fitted line not appear to pass through the data very well?**

<br>
<br>

# Negative Binomial Regression

In this dataset, assuming that the mean is equal to the variance may not be a good idea. Especially considering that the mean count is `r mean(harbordata$Number)` and the variance is `r var(harbordata$Number)`.

Instead, we could consider using a few other methods. In lecture, we discussed the use of quasi-likelihood estimation (Quasi-Poisson). Here, we will talk briefly about the Negative Binomial (NB) distribution.^[While quasi-poisson will produce identical coefficients to using poisson regression, it will result in larger standard errors. In contrast, quasi-poisson will actually produce different coefficients than negative binomial regression. This is because the variance term in quasi-poisson is considered to be a linear function of the mean while the variance term in negative binomial regression is a quadratic (or even cubic) function of the mean. As a result, quasi-poisson and negative binomial regression are more or less appropriate than the other depending on the data scenario.]

In some sense, the Poisson distribution can be thought of as a specific case of the NB distribution where the mean is equal to the variance. By relaxing this assumption though, we end up with two parameters instead of one (the mean and the 'dispersion'). This allows for the scenarios where the variance is much larger than the mean (overdispersion).

**What are some biological scenarios where we might see this?**

<br>
<br>

Below, we will fit a NB model to the dataset and perform the plotting steps as above to see how our results differ.

```{r, fig.cap='**Figure 4: Counts of harbor seals at sites across a range of tide heights (relative to low tide). Solid lines represent model predictions from poisson regression model and negative binomial regression model. Dashed lines represent approximate 95% confidence bands.**'}
library(MASS)

nb1 <- glm.nb(Number ~ substrate + Reltolow, data = harbordata)
summary(nb1)

predDatanb <- data.frame(Reltolow = seq(min(harbordata$Reltolow), max(harbordata$Reltolow), length = 10000), substrate= 'ice')

pred.linknb <- predict(nb1, newdata = predDatanb, se.fit = TRUE)
predDatanb$lambda <- exp(pred.linknb$fit) # exp is the inverse-link function
predDatanb$lower <- exp(pred.linknb$fit - 1.96 * pred.linknb$se.fit)
predDatanb$upper <- exp(pred.linknb$fit + 1.96 * pred.linknb$se.fit)

ggplot() +
  geom_point(data = harbordata, aes(x = Reltolow, y = Number)) +
  geom_path(data = predData, aes(x = Reltolow, y = lambda, color= 'Poisson')) +
  geom_ribbon(data = predData, aes(x = Reltolow, ymin = lower, ymax = upper, color= 'Poisson'),
              fill = NA, linetype= 'dashed') +
  geom_path(data = predDatanb, aes(x = Reltolow, y = lambda, color= 'Negative Binomial')) +
  geom_ribbon(data = predDatanb, aes(x = Reltolow, ymin = lower, ymax = upper, color= 'Negative Binomial'),
              fill = NA, linetype= 'dashed') +
  scale_x_continuous("Tide relative to low (m)") +
  scale_y_continuous("Expected count")
```

<br>

**In what ways do the model predictions from the NB regression differ from the Poisson regression? Compare the information provided by the model output from each model.**

<br>
<br>
<br>


# Assignment

Researchers are interested in understanding factors which affect alligator clutch size (number of eggs). 350 nests were surveyed across 3 habitat types. During the surveys, the clutch size was recorded as well as the length of the female alligator.

The data can be loaded using:

```{r}
library(FANR6750)
data("alligatordata")
```

Create an R markdown report to do the following:

1)  Use both regression methods discussed in lab (Poisson and NB) to estimate clutch size as a function of the available predictors.

2)  Include a well-formatted summary table of the parameter estimates from both models.

3)  Choose one of the models and provide a clear interpretation of the parameter estimates.

4)  Use the poisson regression model to plot the relationship between number of eggs and female length,
    for all three habitat types, on the same graph. The graph should
    include:

    -   the observed data points (color coded by habitat)

    -   a legend

    -   approximate 95% confidence intervals
    
5)  Create a plot of the relationship between number of eggs and female length (at one of the three habitat types) using both regression methods on the same graph. The graph should include:

    -   the observed data points
    
    -   a legend
    
    -   approximate 95% confidence intervals
    
6)  Interpret the plot from Question 5. Which regression method do you think is most appropriate for this dataset? Justify your decision.

As always:

-   Be sure the output type is set to: `output: html_document`

-   Be sure to include your first and last name in the `author` section

-   Be sure to set `echo = TRUE` in all `R` chunks so we can see both
    your code and the output

-   See the R Markdown reference sheet for help with creating `R`
    chunks, equations, tables, etc.

-   See the "Creating publication-quality graphics" reference sheet for
    tips on formatting figures
