---
title: "Schedule (subject to change)"
subtitle: "FANR 6750: Experimental Methods in Forestry and Natural Resources Research"
author: "Fall 2023"
output: 
  rmarkdown::html_vignette:
    keep_md: yes
vignette: >
  %\VignetteIndexEntry{Schedule}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  collapse = TRUE, warning = FALSE, message = FALSE,
  comment = "#>"
)

source(here::here("R/MWF.R"))
```


```{r schedule, echo = FALSE}
library(kableExtra)

schedule <- data.frame(WK = c("1", "", "", 
                              "2", "", "", "",
                              "3", "", "", "", 
                              "4", "", "", "",
                              "5", "", "", "", 
                              "6", "", "", "", 
                              "7", "", "", "", 
                              "8", "", "", "",
                              "9", "", "", "", 
                              "10", "", "", "", 
                              "11", "", "", "", 
                              "12", "", "", "",
                              "13", "", "", "", 
                              "14", "", "", "", 
                              "15", "", "", "", 
                              "16", "", "", "", 
                              "17"),
                       Day = c("W", "F", "", rep(c("M", "W", "F", ""), 15), "M"),
                       Week = MWF(first.day = "2023/08/16", last.day = "2023/12/05"),
                       Lecture = c("Course introduction", "Introduction to statistics", "", #1
      "Introduction to linear models",  "Introduction to linear models", "Principles of statistical inference", "",#2
      "Principles of statiscal inference", "Linear models for simple categorical predictors (aka t-test)", "Linear models for simple categorical predictors (aka t-test)", "", #3
      "Labor Day", "Linear models for simple categorical predictors (aka t-test)", "Null hypothesis testing", "", # 4
       "Null hypothesis testing", "Linear models for categorical predictor w/ >2 levels (aka ANOVA)", "Paper discussion: hypothesis testing", "", #5
       "Linear models for categorical predictor w/ >2 levels (aka ANOVA)", "Linear regression", "Linear regression", "", #6
      "Multiple regression", "Interactions", "buffer", "", #7
       "Diagnosing goodness-of-fit",  "Diagnosing goodness-of-fit", "Model selection",  "", #8 
             "Model selection", "Model selection", "Paper discussion: model selection",  "", #9
      "Random effects", "Nested designs", "Paper discussion: Pseudoreplication", "", # 10
          "Split-plot designs", "Split-plot designs", "Fall break", "", #11
      "Repeated measures", "Repeated measures", "Buffer", "",
         "Regression review", "Analysis of covariance", "Analysis of covariance", "", #13
          "Generalized linear models", "GLM: Logistic regression", "GLM: Logistic regression", "", #14
          "Thanksgiving break", "", "", "", #15
           "GLM: Poisson regression", "GLM: Poisson regression", "GLM: Zero-inflated data",  "", #16
                                   "TBD"),
                       Lab = c("", "", "",
                               "Introduction to R", "", "", "",
                               "Introduction to RMarkdown", "", "", "",
                               "No lab", "", "", "",
                               "R graphics and t-tests", "", "", "",
                               "ANOVA*", "", "", "",
                               "Contrasts and power", "", "", "",
                               "Transformations and nonparametrics*", "", "", "",
                               "Blocking", "", "", "",
                                "Factorial designs", "", "", "",
                               "Nested designs*", "", "", "",
                               "Split-plot designs", "", "", "",
                               "Repeated measures*", "", "", "",
                               "ANCOVA", "", "", "",
                               "Linear models*", "", "", "",
                               "GLMs", "", "", "",
                               # "Model selection", "", "", "",
                               "No lab"),
                       Reading = c("", "Chp. 1", "",
                               "Chp. 2.1-2.3", "", "Chp. 5.2-5.3", "",
                               "", "Chp. 3.1;", "", "",
                               "", "Chp. 8.1-8.4", "", "",
                               "Chp. 8.6", "", "Johnson 1999", "",
                               "Chp. 7.3", "", "", "",
                               "**Exam 1 distributed**", "Chp. 8.5", "", "",
                               "Chp. 8.2; **Exam 1 due**", "Chp. 10.1-10.2", "", "",
                               "Chp. 9.2", "", "Colegrave & Ruxton 2018", "",
                               "", "Chp. 9.1", "", "",
                               "Chp. 11.1-11.3", "", "", "",
                               "**Exam 2 distributed**", "",  "", "",
                               "Chp. 6; **Exam 2 due**", "Chp. 12", "", "",
                               "Chp. 13", "", "", "",
                               "", "", "", "",
                               "**Exam 3 distributed**", "", "Tredennick et al. 2021", "",
                               ""))

schedule %>%
  kable("html", col.names = c("Week", "", "Date", "Lecture topic", "Lab topic", "Reading"), align = 'c') %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = TRUE)

```

\* = Graded assignment

## Tentative lecture topics (aim for 18-20 lectures)

1) Introduction to statistics
    + goals of statistics (exploration, inference, prediction)
    + populations vs samples
    + uncertainty

2) Linear regression
    + model structure: linear predictor + error
    + parameter interpretation
    + assumptions
    
3) Inference
    + sampling distribution
    + standard error
    + confidence intervals

4) Categorical predictor
    + matrix notation 
    + dummy variables
    + formulated as t-test
    
5) Null hypothesis testing
    + p-values
    + power (intuition)

6) Categorical predictor > 2 levels
    + linear model
    + formulated as ANOVA

7) Multiple comparisons
   + contrasts/pre-planned hypotheses

8) Causal inference 

9) Multiple regression
   + continuous + categorical
   + > 1 continuous
   + > categorical
   + link to blocking/ANCOVA

10) Interactions
   + link to factorial design

11) Goodness-of-fit/diagnostics (maybe move earlier)

12) Model selection

13) Random effects/mixed models

14) Nested/split-plot

15) Repeated measures

16) GLM: Logistic regression

17) GLM: Poisson regression

18) GLM: Zero-inflated data
