---
title: "Syllabus"
subtitle: "FANR 6750: Experimental Methods in Forestry and Natural Resources Research"
author: "Fall 2025"
output: 
  rmarkdown::html_vignette:
    keep_md: yes
vignette: >
  %\VignetteIndexEntry{Syllabus}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,warning = FALSE, message = FALSE,
  comment = "#>"
)
```

*Disclaimer*: The course syllabus is a general plan for the course; deviations announced to the class by the instructor may be necessary.

# LOGISTICS

**Lecture**: Monday, Wednesday, Friday 11:30-12:20  
**Location**: 1-304  

**Lab**: Monday 1:50 – 3:45, Tuesday 2:20 – 4:20, or Wednesday 9:10 - 11:10  
**Location**: 4-419  

# INSTRUCTOR
Dr. Clark Rushing  
clark.rushing@uga.edu  
**Office**: Warnell 3-409   
**Office hours**: M 1:00-2:30

# TEACHING ASSISTANTS

Alan Bond (? lab)  
alan.bond1@uga.edu  
**Office**: 1-102  
**Office hours**: M 2:00-4:00 or by appointment  

# COURSE OBJECTIVES

To understand: (1) the logical structure of experiments, including both manipulative and observational experiments, (2) the analysis of such experiments, with a focus on linear models, (3) the use of models in ecological studies (experimental and observational).

# PREREQUISITES

All lab activities for this course will rely heavily on the statistical programming language `R`. During the first few weeks, each lab session will begin with a tutorial on these tools so students are not expected to be experts prior to the course. However, we will quickly move into activities that require some degree of `R` proficiency so I highly recommend that you have a basic understanding of programming in `R` (e.g., importing/exporting & manipulating data objects, visualizing data) prior to this course. If you find yourself struggling with any aspects of using `R`, please seek individual help during office hours. The earlier we can get you up to speed, the more painless the remainder of the semester will be.  


# COURSE FORMAT

The components of a scientific study need to be considered together rather than separately; design and analysis aspects will be completely blended in this course. Emphasis will be on application, and will include instruction on the use of the `R` statistical software. Important points will be reinforced by readings from the scientific literature and by discussions. Homework assignments and exams are intended to keep the material fresh in the students' minds. The syllabus is subject to change, but the order of material is correct. 

# COURSE RESOURCES

## Textbooks

Lecture material will largely be based on:  

**Fieberg, J. (2022). Statistics for Ecologists: A Frequentist and Bayesian Treatment of Modern Regression Models. An open-source online textbook.** 

and 

**Quinn, G.P. & Keough, M.J. 2002. Experimental Design and Data Analysis for Biologists. Cambridge University Press**  

The Fieberg (2022) textbook can be accessed [here](https://statistics4ecologists-v3.netlify.app/). Purchasing the Quinn (2002) textbook is not required but may be helpful for clarifying some concepts. A pdf of the book can be downloaded [here](https://www2.ib.unicamp.br/profs/fsantos/apostilas/Quinn%20&%20Keough.pdf).  

Additional readings will come from a variety of sources, including other textbooks and scientific journals, and will be posted on eLC.

## Lab materials

Materials for labs will be provided as HTML and R Markdown files on the course webpage. These materials will include step-by-step tutorials for all lab exercises as well as links to additional online resources, problem sets, and homework assignments. 

## Software

All lab computers will have `R` and `RStudio` installed on them so students are not required to install software on their own computers. However, students are free to use their own laptops during lab and having `R` and `RStudio` installed will make it easier to complete lab assignments outside of class. Any students wishing to use their own computers **should have R and RStudio installed and running prior to the first lab**. Detailed instructions for installing `R` and `RStudio` can be found [here](http://stat545.com/block000_r-rstudio-install.html).  

If you plan to use your own computer, **be sure you have the most recent versions of both software programs installed**. This will greatly decrease the chances of running into issues running the code I will provide in lab. Prior to the start of the semester, test that you have `R` and `RStudio` installed correctly by doing the following: 

1) Launch `RStudio`

2) Put the cursor in the window labeled `Console`. Type the following code followed by enter or return: `x <- 2 * 4`. Next type `x` followed by enter or return. You should see the value `8` print to screen. If yes, you’ve succeeded in installing R and RStudio.

### Course `R` package

Materials for the course will be distributed through an `R` package called *FANR6750*. The main purpose of the package is to distribute code and data that will used for labs, though eventually additional materials may be included, including lectures and reference documents. You can install the most current version of *FANR6750* with:

``` r
install.packages("devtools")
devtools::install_github("RushingLab/FANR6750")
```

If you encounter any problems with the previous steps, please contact me prior to the first class.  

# COURSE POLICY ON GENERATIVE AI

Focus on this course is on the conceptual basis of modern experimental design and statistical inference, with the primary objective of providing students with the skills and confidence to design, conduct, and report the statistical analyses needed to conduct their own research. For many students, this is the only statistics class they will take in graduate school. For others, this class is only the beginning of progression towards more advanced statistical analyses. In either case, my goal is to provide a firm foundation for conducting and evaluating scientific research.

One of the core philosophies of this course is that there is no one-size-fits-all approach to experimental design and statistical inference. There is no recipe book that students can follow when conducting research and there will often be multiple ways to approach experimental design and data analyses. Successful research therefore requires combining ecological knowledge with thoughtful application of statistical methods. In the complex systems within which we work, critical thinking, domain expertise, and skepticism are paramount. Developing these skills is why you are in graduate school.

Computers are integral to modern statistical inference, allowing researchers to perform complex data formatting, modeling, and visualization tasks that would otherwise be impractical or impossible. As with other resources (textbooks, message boards, blog posts, etc.), AI can help make some tasks (e.g., coding) easier or more efficient when used appropriately. For this reason, the use of AI on homework assignments is not banned in this class (unless specifically noted in the instructions). **However, students must document their use of AI by providing the prompts used and the AI responses for every task for which AI is used.** 

Furthermore, there is increasing evidence that AI may [harm intellectual development](https://thebullshitmachines.com/lesson-11-transforming-education/index.html) if used improperly. In particular, there is a risk that AI *replaces* learning rather than augmenting the learning process. Developing research skills, including statistical analysis, takes practice. Failure, setbacks, and frustration are not only inevitable, they are critical to learning. In my view, the biggest risk of generative AI is that it can provide a shortcut to avoid the short-term challenges associated with learning new concepts or skills. Although that may seem convenient or even helpful, the long term harm of using AI to replace *learning* is detrimental to your development as a researcher, scholar, and citizen. For this reason, exams are structured in a way that AI cannot be used. Some homework questions will also state that AI cannot be used and students are expected to follow these instructions. **Suspected cases of unacknowledged use of AI will be treated as violations of the honor code.** 

# ATTENDANCE

As graduate students who willingly signed up for this course, I presume that you are eager to learn the material and are self-motivated enough to put in the required effort. For this reason, I do not set a formal attendance policy. However, we will cover a lot of material over the course of the semester and each topic will build on concepts from previous weeks. As a result, missing even a few lectures or labs will make it difficult for you to fully master the learning outcomes described above. If you know you will be missing any lectures or labs, please contact me in advance so we can make sure you do not get too far behind on the material. 

### A note on fieldwork

I realize that many students have field work obligations during the semester. If you need to take this course but know in advance that you will be in the field for a portion of the semester, please let me know ASAP so we can discuss whether field work will be a barrier to taking the course or merely an inconvenience. This distinction will mainly be a function of how long and when you will miss class. If these absences are relatively few, taking the course may still be an option. I will do my best to record lectures and make them available via eLC. Students will still be expected to complete and turn in any assignments they miss. If you are going to miss too many classes or will be unable to complete the assignments while you're in the field, it may be better to take the class at a time when your field commitments are smaller. 

# GRADING

200 total points.  

Three exams, 40 points (20%) each. Exams will be completed during normal lecture periods using eLC/Lockdown Browser. Students must complete the exams in person during the scheduled exam period, unless otherwise approved by the instructor at least 1 week before the exam. Remote students must arrange to take exams with a proctor present and exam arrangements must be approved by the instructor prior to the exam. Students will need a laptop with Lockdown Browser installed to complete the exams.

Exams will primarily be focused on concepts, not coding. At least 1 week prior to the exam, students will be given a pool of 8-10 potential questions. The exam itself will consist of 2 questions from this list chosen by the instructor and 2 questions chosen by student. Exam dates on the [schedule](https://rushinglab.github.io/FANR6750/articles/schedule.html) are approximate and subject to change. 

The remaining 80 points (40%) of the grade will come from lab homework assignments. Homework assignments will build on concepts and skills that we cover in lecture and lab. Specific objectives and tasks for each assignment, along with any necessary data, will be provided at the end of each lab activity. These assignments are meant to ensure that you understand lecture and lab concepts, provide additional practice implementing relevant statistical methods in `R`, and provide experience interpreting and presenting model output. 

Over the course of semester, there will be **8** of these assignments worth **10 points each**. Each assignment will be due before class the week after the assignment is posted and must be prepared as an R Markdown file that include all text, code, model output, and figures necessary to fully document your work and conclusions (we will spend the first several labs going over preparation of reports using R Markdown so, again, no previous experience is necessary). You will automatically receive 5 points for turning in a completed lab assignment by the assigned due date. Late assignments will receive a 1-point deduction for each day late and incomplete assignments will have points deducted in proportion to the degree of incompleteness (e.g., if 1/3 of the assignment is incomplete, 2-point deduction). The remaining points will be based on the correctness of your answers. 

All academic work must meet the standards contained in the University's academic [honesty policy](https://honesty.uga.edu/). All students are responsible for informing themselves about those standards before performing any academic work. The penalties for academic dishonesty are [severe](https://honesty.uga.edu/Academic-Honesty-Policy/Consequences_for_Honesty_Violations/), and ignorance is not an acceptable defense.  

If you know ahead of time that you will miss an assignment or exam, please let me know as soon as possible so that we can arrange a make up. If you do not let me or a TA know ahead of time, missed/late assignments will lose 10% of the final grade per day that they are late.

# PRE- AND POST-COURSE ASSESSMENTS

In an effort to help track learning outcomes, students are asked to complete optional pre-course and post-course assessments. Both assessments can be accessed on eLC during the first week (pre-assessment) and last week (post-assessment) of the semester. The assessments are short (~20 questions) and consist primarily of multiple choice and true/false questions that cover a range of topics from throughout the semester, plus several self-assessment questions aimed at gauging students' experience with statistical methods. Responses to these assessments do not count towards the final grade and will be only be used to assess trends in learning outcomes. To incentivize participation, students will receive 2 bonus points for completing the pre-assessment, 2 bonus points for completing the post-assessment, and 1 additional bonus point for completing both (5 bonus points total). 

# GRADING SCALE

```{r grading, echo=FALSE}
library(kableExtra)
grades <- data.frame(Grade = c("A", "A-", "B+", "B", "B-", "C+", "C", "C-", "D+", "D", "F"),
                       Percent = c("93-100%", "90-92.9%", 
                                   "87-89.9%", "83-86.9%", "80-82.9%", 
                                   "77-79.9%", "73-76.9%", "70-72.9%",
                                   "67-69.9%", "60-66.9%", "59.9% and below"),
                       Points = c("233-250", "225-232.5", 
                                   "217.5-224.5", "207.5-217.5", "200-207.5", 
                                   "192.5-200", "182.5-192.5", "175-182.5",
                                   "167.5-175", "150-167.5", "0-150"))

grades %>%
  kable("html", col.names = c("GRADE", "% RANGE", "POINT RANGE"), align = 'c') %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = TRUE)

```

# UNIVERSITY HONOR CODE & ACADEMIC HONESTY

As a University of Georgia student, you have agreed to abide by the UGA academic honesty policy. UGA Student Honor code: 

> I will be academically honest in all of my academic work and will not tolerate academic dishonesty of others  

A Culture of Honesty, the University's policy and procedures for handling cases of suspected dishonesty, can be found at [https://honesty.uga.edu/](https://honesty.uga.edu/)  

You are responsible for informing yourself about the university’s standards before performing any academic work. Lack of knowledge of the academic honesty policy is not a reasonable explanation for a violation. Please ask if you have questions related to course assignments and the academic honesty policy. Any form of possible academic dishonesty will be reported to the UGA Office of the Vice President for Instruction.

# ACCOMMODATIONS FOR DISABILITIES

If you require a disability-required accommodation, it is essential that you register with the Disability Resource Center (Clark Howell Hall; https://drc.uga.edu; 706-542-8719 [voice]; 706-542-8778 [TTY]) and notify me of your eligibility for reasonable accommodations. We can then plan how best to coordinate your accommodations. Please note that accommodations cannot be provided retroactively. 

# WELLNESS STATEMENT 

Mental Health and Wellness Resources:

- If you or someone you know needs assistance, you are encouraged to contact Student Care and Outreach in the Division of Student Affairs at 706-542-7774 or visit [https://sco.uga.edu/](https://sco.uga.edu/). They will help you navigate any difficult circumstances you may be facing by connecting you with the appropriate resources or services.

- UGA has several resources for a student seeking mental health services ([https://www.uhs.uga.edu/bewelluga/bewelluga](https://www.uhs.uga.edu/bewelluga/bewelluga)) or crisis support ([https://www.uhs.uga.edu/info/emergencies](https://www.uhs.uga.edu/info/emergencies)).

- If you need help managing stress anxiety, relationships, etc., please visit BeWellUGA ([https://www.uhs.uga.edu/bewelluga/bewelluga](https://www.uhs.uga.edu/bewelluga/bewelluga)) for a list of FREE workshops, classes, mentoring, and health coaching led by licensed clinicians and health educators in the University Health Center.

- Additional resources can be accessed through the UGA App.


# FERPA NOTICE

The Federal Family Educational Rights and Privacy Act (FERPA) grants students certain information privacy rights. To comply with FERPA, all communication that refers to individual students must be through a secure medium (UGAMail or eLC) or in person. Instructors are not allowed to respond to messages that refer to individual students or student progress in the course through non-UGA accounts, phone calls, or other types of electronic media. For details, please visit [https://apps.reg.uga.edu/FERPA](https://apps.reg.uga.edu/FERPA).
