---
format:
  revealjs:
    highlight-style: github
    incremental: false
    theme: FANR6750.scss
    reference-location: document
knitr:
  opts_chunk:
    dev: png
    dev.args: { bg: "transparent" }
from: markdown+emoji
---

```{r setup}
#| echo: false
#| include: false
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo = FALSE, fig.align = 'center', warning = FALSE,
                      message = FALSE, fig.retina = 2)
source(here::here("R/zzz.R"))
library(FANR6750)
```

## LECTURE 5: null hypothesis testing {.theme-title .center}

::: footer
FANR 6750  
Fall 2026
:::

## Outline {.theme-inverse .smaller}

<br/>

### 1) Sampling error and regression coefficients

<br/>  

::: {.fragment}
### 2) Logic of null hypothesis testing

<br/> 
:::

::: {.fragment}
### 3) *t*-statistics and *t*-distribution

<br/> 
:::

::: {.fragment}
### 4) *p*-values and decisions

<br/> 
:::

::: {.fragment}
### 5) NHT for regression models

<br/> 
:::

## Fitting the model in `R` {.smaller}

```{r echo=TRUE}
data(frogdata)

fit2 <- lm(Frogs ~ Development, data = frogdata)

broom::tidy(fit2)
```


- The `(Intercept)` estimate ($\beta_0$) is the expected number of frogs in a low-development plot


- The `DevelopmentHigh` estimate ($\beta_1$) is the *difference* between low-development and high-development plots


- [Question]{.highlight}: Are there fewer frogs in high-development plots?

## Sampling (again) {.smaller}

[Question]{.highlight}: Are there fewer frogs in high-development plots?

- This question is harder to answer than it might seem. Why?

:::{.fragment}

Sampling error!

:::

:::{.incremental}
- If we did the experiment again, we'd get a different difference (maybe even a change in direction)

- The sample means will *always* be different (which we will prove shortly)

- If the sample means will never be the same, we need to re-formulate our question:

    + How large of a difference between our *samples* provides evidence that the *population* means are different? 

:::

## Null hypothesis testing {.theme-inverse .center}


## Null hypothesis testing {.smaller}

How can we express our question about development and its impacts on frogs as a hypothesis?

:::{.fragment}

If development does not influence frog abundance, $\beta_1 = 0$

- This is called the *null hypothesis* (i.e., no effect)

- Denoted $H_0$

:::

:::{.fragment}

If development does influence frog abundance, $\beta_1 \neq 0$

- This is the *alternative hypothesis*

- Denoted $H_a$

:::

:::{.fragment}

- Note that we could make the alternative more specific and say development negatively effects abundance ($\beta_1 < 0$)

:::

## Null hypothesis testing {.smaller}

Remember that $\beta_1$ is a [population]{.highlight} parameter

- We estimate population parameters from samples

:::{.fragment}

What is our best estimate of $\beta_1$? 

:::

:::{.fragment}
 - The difference in sample means: $\hat{\beta}_1 = \hat{\mu}_{Low} - \hat{\mu}_{High}$
 
:::

:::{.fragment}

- For the frog data:

```{r echo = TRUE}
mu_high <- mean(frogdata$Frogs[frogdata$Development == "High"])
mu_low <- mean(frogdata$Frogs[frogdata$Development == "Low"])

(beta_1.hat <- mu_high - mu_low)
```

Note that this is same as the estimate of $\beta_1$ provided by the `lm` and `t-test` functions

:::


## Null hypothesis testing {.smaller}

Our estimate of $\beta_1$ is clearly not $0$ but does that imply that the null hypothesis is wrong?

:::{.incremental}

- Not necessarily! Why?


- Sampling error! Even if $\beta_1 = 0$, $\hat{\beta}_1$ will *never* equal 0

:::

:::{.fragment}

Enter *null hypothesis testing*:

> formal approach to deciding whether a statistical relationship in a sample reflects a real relationship in the population or is just due to chance 

:::


## Null hypothesis testing {.smaller}

NHT is based on the *expectation* of what our data should look like [if the null hypothesis is true]{.highlight}

:::{.incremental}

- If our data look really different than what we expect [if the null hypothesis is true]{.highlight}, then it is unlikely that the null hypothesis is true and we reject $H_0$


- It's important to note that there is *always* a chance that our results are simply due to chance 

:::

:::{.fragment}

Type I error (i.e., false positive rate)

- The probability that we will reject $H_0$ when it is actually true

- Usually denoted $\alpha$

- Generally, $\alpha = 0.05$ or $\alpha = 0.01$ are accepted Type I error rates

:::


## Null hypothesis testing {.smaller}

How do we know what the data should look like [if the null hypothesis is true]{.highlight}?

:::{.incremental}

- Again, we can use theory to tell us about long-term expectations (similar to the sampling distribution)


- But maybe easier to understand using simulation

:::


## Simulating data under $\small H_0$ {.smaller}

Imagine the null hypothesis is true in the measurement error example (i.e., measurement error = 0)

- Any difference between the sample mean and 0 is just due to sampling error

:::{.fragment}

- We can simulate a "sample" in `R` using the `rnorm()` function

```{r echo = TRUE}
y <- rnorm(n = 10, mean = 0, sd = 1)

mean(y) - 0
```


:::

:::{.fragment}

Note that the we fixed the mean of the normal distribution to be 0, so we know the null hypothesis is true

:::


## Simulating data under $\small H_0$ {.smaller}

Instead of just taking the difference $\bar{y} - 0$, we can standardize the differences by dividing by the standard error:

```{r echo = TRUE}
se.y <- sd(y)/sqrt(10)
(mean(y) - 0)/se.y
```

:::{.fragment}

- This tells us how many standard errors away from 0 the sample mean is 

- We'll call this value $t$

$$ t = \frac{\bar{y} - 0}{SE_y}$$

:::


## Simulating data under $\small H_0$ {.smaller}

Generate and plot 1000 sample means:

```{r echo = TRUE}
t <- numeric(length = 1000)

for(i in 1:1000){
  y <-  rnorm(10, mean = 0, sd = 1)
  t[i] <- mean(y)/(sd(y)/sqrt(10))
}
```

```{r fig.width=6, fig.height=2.5}
ggplot(data.frame(t = t), aes(x = t)) +
  geom_histogram(color = "grey75") +
  geom_rug(alpha = 0.1) +
  scale_x_continuous("t") +
  scale_y_continuous("Count")
```


## Simulating data under $\small H_0$ {.smaller}

```{r fig.width=6, fig.height=3}
ggplot(data.frame(t = t), aes(x = t)) +
  geom_histogram(color = "grey75") +
  geom_rug(alpha = 0.1) +
  scale_x_continuous("t") +
  scale_y_continuous("Count")
```

Note that:

:::{.incremental}
1. The null hypothesis was true for every sample


2. Some sample means were bigger than expected, some were smaller (all due to sampling error!)


3. The resulting distribution looks *kind of* normal

:::


## The *t*-distribution {.smaller}

The distribution of the *t*-statistics is not quite normally distributed

:::{.fragment}

Instead, theory says that the *t*-statistics will follow a $t$-distribution with $n - 1$ degrees of freedom ([if the null hypothesis is true]{.highlight}!)

```{r t, fig.width=8, fig.height=6}
#| fig-width: 8
#| fig-height: 5

curve(dt(x, 1), -5, 5, ylim=c(0, 0.5),
      xlab="t value", ylab="Probability density",
      cex.lab=1.5, lwd=1.5)
curve(dt(x, 5), -5, 5, add=TRUE, col="blue", lwd=1.5)
curve(dnorm(x, 0, 1), -5, 5, add=TRUE,  lty=3, lwd=1.5)
legend(-5, 0.5, c("Standard normal distribution", "t distribution (df=5)", "t distribution (df=1)"),
       lty=c(3, 1, 1), col=c("black", "blue", "black"), lwd=1.5)
```

:::

## The *t*-distribution {.smaller}

More about the *t*-distribution

:::{.incremental}

- Continuous probability distribution

- Symmetrical with mean = 0

- More mass in the tails as degrees of freedom get smaller (i.e., more extreme values become more likely)

- For sample sizes $n \gt 30$, the $t$-distribution is essentially a standard normal distribution with mean = 0 and SD = 1

- Published by William Sealy Gosset in 1908 under the pseudonym "Student". Gosset worked for Guinness and was interested in quality control of beer ingredients 

:::

## Null hypothesis testing {.smaller}

[Quick review]{.highlight}:

:::{.incremental}

1. The null hypothesis $H_0$ is that there is no effect or no difference


2. The *t*-statistic measures the difference between the sample mean and its hypothesized value (under the null) relative to its standard error 


3. [If the null hypothesis is true]{.highlight}, the *t*-statistics from repeated samples follow a *t*-distribution with $n-1$ degrees of freedom

:::

:::{.fragment}
Importantly, because we can quantify properties of the *t*-distribution, we can compare the *t*-statistic calculated from our observed sample to the expected values under the null hypothesis

:::{.incremental}

- If our observed *t*-statistic would be unlikely under the null hypothesis, we can conclude that the null hypothesis is false


- This is the logic behind null hypothesis testing

:::

:::

## Null hypothesis testing {.smaller}

Is there evidence to reject $H_0$?

```{r fig.width=6, fig.height=4}
df <- data.frame(t = t,
                 group = ifelse(t > -1.07 & t < 1.07, "0", "1"))

ggplot() +
  geom_histogram(data = df, aes(x = t), 
                 alpha = 0.75, bins = 100) +
  geom_vline(xintercept = 1.05, color = "#D47500", linetype = "dashed", size = 1) +
  geom_vline(xintercept = mean(t), color = "black") +
  scale_y_continuous("") + theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())
```

## Null hypothesis testing {.smaller}

Measurement error example

```{r echo =TRUE}
y <- c(-0.062, -0.38, 0.85, -0.58,  0.53,  0.09,  0.31,  0.77,  0.59,  -0.17)

(mean.y <- mean(y))

(se.y <- sd(y)/sqrt(10))

(t.stat <- mean.y / se.y)
```

:::{.fragment}

Is there evidence to reject $H_0$?

:::

## Null hypothesis testing {.smaller}

Measurement error example

```{r}
#| fig-width: 6
#| fig-height: 4
df <- data.frame(t = t,
                 group = ifelse(t > -t.stat & t < t.stat, "0", "1"))

ggplot() +
  geom_histogram(data = df, aes(x = t, fill = as.factor(group)), 
                 alpha = 0.75, bins = 100) +
  geom_vline(xintercept = t.stat, color = "#D47500", linetype = "dashed", size = 1) +
  geom_vline(xintercept = mean(t), color = "black") +
  scale_y_continuous("") + theme(axis.text.y = element_blank(), axis.ticks.y = element_blank()) +
  guides(fill = "none") +
  labs(subtitle = "Distribution of 1000 t-statistics")
```

Approximately 18% of the simulated values of $t$ are larger than 1.25 (or smaller than -1.25)

- Put another way, **if the null hypothesis is true**, there is about a 1 in 5 chance of observing $t \geq 1.25$ 


## Null hypothesis testing {.smaller}

In reality, we don't need to simulate the distribution of $t$ every time we do an experiment

<br/>
```{r p}
#| fig-width: 5
#| fig-height: 3
op <- par(mai=c(0.8, 0.2, 0.2, 0.2))
curve(dt(x, df=9), -4, 4, xlab="t value", ylab="", yaxt="n",
      ylim=c(0,0.5),
      frame=FALSE , main="t-distribution with df=9", cex.main=1.5)
xs1 <- seq(qt(.025, df=9), -4, by=-0.1)
ys1 <- dt(xs1, df=9)
xs2 <- seq(qt(.975, df=9), 4, by=0.1)
ys2 <- dt(xs2, df=9)
xs3 <- seq(qt(pt(-1.251,9), df=9), -4, by=-0.1)
ys3 <- dt(xs3, df=9)
xs4 <- seq(qt(pt(1.251,9), df=9), 4, by=0.1)
ys4 <- dt(xs4, df=9)
polygon(c(xs3, rev(xs3)), c(rep(0, length(xs3)), rev(ys3)), col=gray(0.7))
polygon(c(xs4, rev(xs4)), c(rep(0, length(xs4)), rev(ys4)), col=gray(0.7))
#polygon(c(xs1, rev(xs1)), c(rep(0, length(xs1)), rev(ys1)), col="red")
#polygon(c(xs2, rev(xs2)), c(rep(0, length(xs2)), rev(ys2)), col="red")
text(-1.251, dt(0,9), "t=-1.251", pos=3)
text(1.251, dt(0,9), "t=1.251", pos=3)
#text(-3, dt(0,18), "t=-3", pos=3)
#text(3, dt(0,18), "t=3", pos=3)
arrows(-1.07, dt(0,9), -1.251, dt(-1.07,9), length=0.1)
arrows(1.07, dt(0,9), 1.251, dt(1.07,9), length=0.1)
par(op)
```


- Easy (in `R`) to calculate the area of grey shaded regions, which is the probability of getting a value $t$ larger in magnitude than 1.251 **if the null hypothesis true**


- This is called a *p*-value


- In our example, $p = 0.174$. Would you reject the null?


## More on *p*-values {.smaller}

The *p* -value tells you how likely your observations (or more extreme) would be [if the null hypothesis is true]{.highlight}  

:::{.fragment}

A *p* -value does not tell us how much evidence there is in favor of a particular difference in means

:::

:::{.fragment}

What factors result in a small *p* -value?

:::

:::{.incremental}

- The sample mean is far from 0

- And/or the SE is small
:::


## Type i error and *p*-values {.smaller}

In NHT, our conclusion must be to either reject or "fail to reject" the null hypothesis

:::{.fragment}

When we reject the null hypothesis, there is always a chance that we do so mistakenly

- Due to sampling error, there is always a chance we get a large value of *t* even if the null hypothesis is true

- These "false positive" mistakes are referred to as *Type I error* (denoted $\alpha$)

:::

:::{.fragment}

The probability of type I error is measured by the *p*-value

:::

:::{.fragment}

Generally, we want to avoid false positive conclusions. Why?

:::

:::{.fragment}

- Type I error rates of $\lt 5$% or $\lt 1$% are generally considered reasonable

:::

## Critical values {.smaller}

Before statistical software made it easy to calculate *p*-values, researchers would look up *critical values*

- For a given sample size (degrees of freedom) and $\alpha$, what is the associated value of *t*

- If your calculated *t* is $\geq$ the critical value, reject the null hypothesis

```{r cv, fig.width=6, fig.height=4}
op <- par(mai=c(0.8, 0.2, 0.2, 0.2))
curve(dt(x, df=9), -4, 4, xlab="t value", ylab="", yaxt="n",
      ylim=c(0,0.5),
      frame=FALSE , main="t distribution with df=9", cex.main=1.5)
xs1 <- seq(qt(.025, df=9), -4, by=-0.1)
ys1 <- dt(xs1, df=9)
xs2 <- seq(qt(.975, df=9), 4, by=0.1)
ys2 <- dt(xs2, df=9)
polygon(c(xs1, rev(xs1)), c(rep(0, length(xs1)), rev(ys1)), col="red")
polygon(c(xs2, rev(xs2)), c(rep(0, length(xs2)), rev(ys2)), col="red")
text(xs1[1], dt(0,9)/1.5, "critical value\nt=-2.26", pos=3)
text(xs2[1], dt(0,9)/1.5, "critical value\nt=2.26", pos=3)
arrows(xs1[1], dt(0,9)/1.5, xs1[1], ys1[1], length=0.1)
arrows(xs2[1], dt(0,9)/1.5, xs2[1], ys2[1], length=0.1)
#text(-3, dt(0,18), "t=-3", pos=3)
#text(3, dt(0,18), "t=3", pos=3)
#arrows(-3, dt(0,18), -3, dt(-3,18), length=0.1)
#arrows(3, dt(0,18), 3, dt(3,18), length=0.1)
par(op)

```


## Frog example {.smaller}

Back to the frog example

```{r echo=TRUE}
data(frogdata)

fit2 <- lm(Frogs ~ Development, data = frogdata)

broom::tidy(fit2)
```


Now we're ready to interpret the rest of this output


## Nht for regression parameters {.smaller}

It turns out, regression parameters also follow a *t*-distribution **if the null hypothesis is true**

:::{.fragment}

In a regression context:

$$ t = \frac{\hat{\beta}_1 - \beta_1}{SE(\hat{\beta}_1)}$$

:::

:::{.fragment}

For the frog example:

```{r echo = TRUE}
beta1.hat <- mu_high - mu_low
(t <- (beta1.hat - 0)/2.64)
```

:::


## Nht for regression parameters {.smaller}

```{r echo=TRUE}
broom::tidy(fit2)
```

Notice that the `lm` output provides a *t*-statistic and *p*-value for both the slope coefficient and the intercept

- How do we interpret the intercept *p*-value? 

- Is the *p*-value for the intercept biologically meaningful?



## Nst for t-tests {.smaller}

As we saw before, the frog model can be fit as a two-sample *t*-test

```{r echo = TRUE}
t.test(Frogs ~ Development, data = frogdata)
```


Notice that the *t*-statistic and *p*-value is the same as the `lm` model


## Looking ahead

<br/>

[Next time]{.highlight}: Evaluating assumptions of linear models

<br/>

[Reading]{.highlight}: [Fieberg 1.4-1.5](https://statistics4ecologists-v3.netlify.app/01-linearregression#sec-lmassumptions)
